---
title: "Right to life"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
execute:
  warning: false
  message: false
  out-width: 100%
---

```{r setup}
#| echo: false
library(tidyverse)
library(cowplot)
library(rvest)
library(ggbeeswarm)
library(scales)
library(httr)
library(jsonlite)
library(ggrepel)
library(readxl)
library(ggmosaic)
library(here)

source(here("consts.R"))

dir.create(here("drafts/rightToLife/datasets"), recursive = TRUE, showWarnings = FALSE)
path <- here("drafts/rightToLife/datasets/")
raw <- here("data/raw/")
```

# Human Rights Defenders

```{r global-hrd-deaths}
world <- c(426, 423, 416, 487, 437, 387, 401, 500, 595, 685, 743)
years <- 2015:2025

hrds_annual_export <- data.frame(world, years)

hrds_annual_export %>%
    write.csv(paste0(path, "/hr_defenders_killed_globally.csv"), row.names = FALSE)

hrds_annual_export %>%
    ggplot(aes(x = years, y = world)) +
    geom_col(fill = orange) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x = "Year", y = "Number of human rights defenders killed annually") +
    theme_minimal_hgrid()
```

```{r regional-hrd-beeswarm}
#| fig-height: 7
cols <- c("Category", "x", "Central Asia and Southern Asia", "Eastern Asia and South-eastern Asia", "Latin America and the Caribbean", "Northern America and Europe", "Sub-Saharan Africa", "Western Asia and Northern Africa", "World", "Min", "Max", "start", "end")

wide <- read_csv(I("Women HRDs,1,12%,19%,16%,19%,7%,11%,14%,7%,19%,7%,12%
Environment & land HRDs,2,12%,25%,52%,6%,19%,1%,31%,1%,52%,1%,51%
Indigenous & minority HRDs,3,18%,14%,45%,13%,18%,13%,29%,13%,45%,13%,32%
Youth HRDs,4,8%,6%,10%,31%,10%,13%,10%,6%,31%,6%,25%"),
    col_names = cols
) %>%
    select(-x, -World, -Min, -Max, -start, -end) %>%
    mutate(across(!Category, ~ as.numeric(str_extract(., "[0-9]*"))))

hrds_category_export <- wide %>%
    pivot_longer(-Category) %>%
    mutate(
        Category = factor(Category),
        cat = as.numeric(Category)
    )

hrds_category_export %>%
    write.csv(paste0(path, "/hr_defenders_regional_category.csv"), row.names = FALSE)

hrds_category_export %>%
    ggplot(aes(x = str_wrap(as.character(Category), 20), y = value / 100, color = name)) +
    geom_beeswarm(method = "center", cex = 2, size = 3) +
    scale_color_manual(values = palettes$cats) +
    labs(x = "Category", y = "Percentage of total killed or disappeared", color = "Region") +
    scale_y_continuous(
        expand = c(0, 0),
        limits = c(0, .55),
        labels = label_percent()
    ) +
    theme_minimal_hgrid() +
    theme(legend.position = "bottom") +
    guides(color = guide_legend(nrow = 3, byrow = TRUE))
```

# Conflict-related deaths

```{r global-crds}
crd_files <- list.files(paste0(raw, "/crd"))

l <- list()

for (f in crd_files) {
    l[[str_remove(f, ".rds")]] <- readRDS(paste0(raw, "/crd/", f))
}

crds_total_export <- l$series_total_w

crds_total_export %>% write.csv(paste0(path, "/global_crds_annual.csv"), row.names = FALSE)

crds_total_export %>%
    ggplot(aes(x = as.factor(year), y = deaths_OHCHR)) +
    geom_line(color = orange, group = 1) +
    scale_y_continuous(
        limits = c(0, max(crds_total_export$deaths_OHCHR)),
        expand = c(0, 0),
        labels = label_comma()
    ) +
    labs(x = "Year", y = "Number of global conflict-related deaths") +
    theme_minimal_grid()
```

# Life expectancy

```{r regional-lifeexp}
library(rvest)
lifeexp <- read.csv(paste0(raw, "/lifeexp_who.csv"))

page <- read_html("https://unstats.un.org/unsd/methodology/m49/overview/")
t <- html_element(page, "table")
unregions <- html_table(t, header = TRUE) %>% select(`Region Name`, `M49 Code`, `Sub-region Name`, `ISO-alpha3 Code`, `Country or Area`)

lifeexp_export <- lifeexp %>%
    left_join(unregions, by = c("SpatialDimValueCode" = "ISO-alpha3 Code")) %>%
    filter(Period == 2021, Dim1ValueCode == "SEX_BTSX", Indicator == "Life expectancy at birth (years)") %>%
    select(Location, FactValueNumeric, `Region Name`)

write.csv(lifeexp_export, paste0(path, "/life_expectancy_beeswarm.csv"), row.names = FALSE)

lifeexp_export %>%
    ggplot(aes(x = FactValueNumeric, y = `Region Name`, color = `Region Name`)) +
    geom_beeswarm(cex = 2) +
    scale_color_manual(values = palettes$regions, guide = "none") +
    scale_x_continuous(expand = c(0, 0), limits = c(50, 85)) +
    theme_minimal_vgrid() +
    labs(color = "Region", y = NULL, x = "Life expectancy at birth")
```

# Homicides

```{r homicide-range}
body <- list(
    limit = 1000,
    where = list(
        datetime = c("2000-01-01", "2024-01-01"),
        sexCode = "_T"
    )
)

response <- POST(
    url = "https://hrdx-api-staging.un.org/vcIhrPsrc/search",
    body = toJSON(body, auto_unbox = TRUE),
    content_type_json()
)

homicide_per_100k <- fromJSON(content(response, "text"), flatten = TRUE)$results

body <- list(
    limit = 1000
)

response <- POST(
    url = "https://hrdx-api-staging.un.org/region/search",
    body = toJSON(body, auto_unbox = TRUE),
    content_type_json()
)

all_regions <- fromJSON(content(response, "text"), flatten = TRUE)$results %>% select(id, name, level, parentId)

regions <- fromJSON(content(response, "text"), flatten = TRUE)$results %>%
    select(id, name, level) %>%
    filter(level %in% 0:1)

homicide_per_100k_export <- homicide_per_100k %>%
    left_join(regions, by = c("regionId" = "id")) %>%
    filter(!is.na(name)) %>%
    mutate(
        year = year(datetime),
        highlight = ifelse(name == "World", "Y", "N")
    ) %>%
    select(name, year, value, highlight)

homicide_per_100k_export %>%
    pivot_wider(names_from = year, values_from = value) %>%
    mutate(incdec = ifelse(`2024` - `2000` > 0, "increased", "decreased")) %>%
    write.csv(paste0(path, "/homicide_100k_range.csv"), row.names = FALSE)

label_2000 <- homicide_per_100k_export %>%
    filter(year == 2000) %>%
    mutate(label_x = 1999)

label_2023 <- homicide_per_100k_export %>%
    filter(year == 2024) %>%
    mutate(label_x = 2025)

label_data <- data.frame(x = c(2000, 2024), label = c(2000, 2024), y = 0)

ggplot(homicide_per_100k_export, aes(x = year, y = round(value, 1), group = name)) +
    geom_line(aes(color = highlight, alpha = highlight)) +
    scale_alpha_manual(values = c("Y" = 1, "N" = .9)) +
    geom_text_repel(
        data = label_2000,
        aes(x = label_x, y = value, label = paste0(name, "\n", value), color = highlight),
        direction = "y",
        hjust = 1,
        segment.color = NA,
        xlim = c(-Inf, 1999.75),
        lineheight = 0.8,
        size = 3
    ) +
    geom_text_repel(
        data = label_2023,
        aes(x = label_x, y = value, label = paste0(name, "\n", value), color = highlight),
        direction = "y",
        hjust = 0,
        segment.color = NA,
        xlim = c(2023.25, Inf),
        lineheight = 0.8,
        size = 3
    ) +
    geom_segment(aes(x = 2000, xend = 2000, y = .5, yend = 16), inherit.aes = FALSE) +
    geom_segment(aes(x = 2024, xend = 2024, y = .5, yend = 16), inherit.aes = FALSE) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = palettes$highlight) +
    scale_x_continuous(expand = expansion(mult = c(0.3, 0.3))) +
    geom_text(data = label_data, aes(x = x, y = y, label = label), inherit.aes = FALSE) +
    theme_void() +
    guides(color = "none", label = "none", alpha = "none")
```

```{r homicide-snapshot}
body <- list(
    limit = 1000,
    where = list(
        sexCode = "_T",
        datetime = "2024-01-01"
    )
)

response <- POST(
    url = "https://hrdx-api-staging.un.org/vcIhrPsrcn/search",
    body = toJSON(body, auto_unbox = TRUE),
    content_type_json()
)

homicide_n_export <- fromJSON(content(response, "text"), flatten = TRUE)$results %>%
    mutate(
        countryId = as.integer(countryId),
        year = year(datetime)
    ) %>%
    left_join(unregions, by = c("countryId" = "M49 Code")) %>%
    group_by(`Region Name`) %>%
    summarize(total = sum(value), .groups = "drop") %>%
    mutate(
        value = total / sum(total) * 100,
        n = n()
    )

homicide_n_export %>%
    write.csv(paste0(path, "/homicide_n_2024.csv"), row.names = FALSE)

homicide_n_export %>%
    arrange(total) %>%
    mutate(
        csum = rev(cumsum(rev(value))),
        pos = value / 2 + lead(csum, 1),
        pos = if_else(is.na(pos), value / 2, pos)
    ) %>%
    ggplot(aes(x = "", y = value, fill = fct_inorder(`Region Name`))) +
    geom_col(width = 1, color = 1) +
    coord_polar(theta = "y") +
    scale_fill_manual(values = palettes$regions, guide = "none") +
    geom_text_repel(
        aes(y = pos, label = paste0(`Region Name`, "\n", total, " homicides")),
        size = 4.5, nudge_x = .8, show.legend = FALSE
    ) +
    theme_void()
```

```{r homicide-sex-scatter}
#| fig-height: 5
body <- list(
    limit = 1000,
    where = list(
        datetime = c("2024-01-01"),
        sexCode = c("M", "F")
    )
)

response <- POST(
    url = "https://hrdx-api-staging.un.org/vcIhrPsrc/search",
    body = toJSON(body, auto_unbox = TRUE),
    content_type_json()
)

homicides_sex <- fromJSON(content(response, "text"), flatten = TRUE)$results

homicides_sex_export <- homicides_sex %>%
    mutate(
        countryId = as.integer(countryId),
        year = year(datetime)
    ) %>%
    filter(year == 2024, !is.na(countryId)) %>%
    left_join(unregions, by = c("countryId" = "M49 Code")) %>%
    select(`Country or Area`, `Region Name`, value, year, sexCode) %>%
    pivot_wider(names_from = sexCode, values_from = value)

homicides_sex_export %>%
    write.csv(paste0(path, "/homicides_sex.csv"), row.names = FALSE)

homicides_sex_export %>%
    ggplot(aes(x = M, y = F)) +
    geom_point(aes(color = `Region Name`)) +
    scale_color_manual(values = palettes$region) +
    theme_minimal_grid(12) +
    labs(x = "Intentional homicides of men, per 100,000 population", y = "Intentional homicides of women, per 100,000 population") +
    scale_x_continuous(expand = c(.02, 0)) +
    scale_y_continuous(expand = c(.02, 0))
```

Data obtained at UNODC: <https://data.unodc.org/datareport/hom-victim>

```{r ipv}
#| fig-height: 6
# obtained here: https://data.unodc.org/datareport/hom-victim
homicides_unodc <- read_excel(paste0(raw, "/unodc_homicide.xlsx"), skip = 2)

homicides_ipv_export <- homicides_unodc %>%
    filter(
        Dimension == "by relationship to perpetrator",
        `Unit of measurement` == "Rate per 100,000 population",
        Sex %in% c("Male", "Female"),
        Year == 2022,
        Category %in% c(
            "Perpetrator unknown to the victim",
            "Intimate partner or family member: Intimate partner",
            "Intimate partner or family member: Family member"
        )
    ) %>%
    group_by(Sex, Category) %>%
    summarize(value = mean(VALUE)) %>%
    pivot_wider(names_from = Sex, values_from = value) %>%
    mutate(
        diff = Female - Male,
        sign = ifelse(diff > 0, "Female", "Male"),
        Category = str_to_lower(str_remove(Category, "Intimate partner or family member: |Perpetrator "))
    )

homicides_ipv_export %>%
    write.csv(paste0(path, "/homicides_ipv.csv"), row.names = FALSE)

homicides_ipv_export %>%
    ggplot(aes(x = Category, y = diff, fill = sign)) +
    geom_col() +
    scale_fill_manual(values = palettes$gender, guide = "none") +
    labs(x = "Relationship to victim", y = "Difference in rate of intentional homicide, Female - Male") +
    geom_hline(yintercept = 0) +
    theme_minimal_hgrid()
```

# Maternal mortality

The MMR rose in 2021 (from 282 000 maternal deaths in 2020 to 322 000 maternal deaths in 2021). It is possible that the COVID-19 pandemic contributed to this via two mechanisms: (i) indirect obstetric deaths – where the woman had SARS-CoV-2 infection and died as a result of the interaction between COVID-19 and her pregnant state; and/or (ii) direct obstetric deaths – where disruptions to health services hindered access to and/or quality of care resulting in pregnancy complications that ended in mortality, when they may otherwise have been prevented or managed successfully (*4)*. The data show that this interruption to the downward trajectory of global MMR was short-lived. In 2022, the global MMR and number of maternal deaths were lower than they had been in the three years immediately prior to the COVID-19 pandemic. 

<https://www.who.int/news-room/fact-sheets/detail/maternal-mortality>

```{r maternal-mort}
body <- list(
    limit = 1000,
    where = list("regionId" = c(19, 2, 9, 142, 150))
)

response <- POST(
    url = "https://hrdx-api-staging.un.org/shStaMort/search",
    body = toJSON(body, auto_unbox = TRUE),
    content_type_json()
)

maternal_mort_export <- fromJSON(content(response, "text"), flatten = TRUE)$results %>%
    inner_join(regions, by = c("regionId" = "id")) %>%
    mutate(year = year(datetime)) %>%
    select(year, name, value)

maternal_mort_export %>%
    ggplot(aes(x = year, y = value, color = name)) +
    geom_line() +
    scale_color_manual(values = palettes$regions) +
    theme_minimal_grid() +
    labs(y = "Number of maternal deaths per 100,000 live births", x = "Year", color = "Region") +
    geom_vline(xintercept = 2021, linetype = "dashed") +
    geom_text(x = 2016, y = 650, label = "COVID-19 pandemic", color = "black")
```

# Under-5 mortality rate

```{r under-5}
body <- list(
    limit = 1000,
    where = list(
        "datetime" = "2024-01-01",
        "sexCode" = "_T"
    )
)

response <- POST(
    url = "https://hrdx-api-staging.un.org/shDynMort/search",
    body = toJSON(body, auto_unbox = TRUE),
    content_type_json()
)

under5_mort_export <- fromJSON(content(response, "text"), flatten = TRUE)$results %>%
    filter(!is.na(regionId)) %>%
    left_join(all_regions, by = c("regionId" = "id")) %>%
    mutate(parentId = ifelse(level == 1, regionId, parentId)) %>%
    filter(name != "Australia and New Zealand") %>%
    select(-name, -level) %>%
    left_join(all_regions, by = c("parentId" = "id")) %>%
    group_by(name) %>%
    summarize(mean_regional = mean(value))

under5_mort_export %>%
    write.csv(paste0(path, "/under5-mortalities2024.csv"), row.names = FALSE)

under5_mort_export %>%
    ggplot(aes(x = fct_reorder(name, -mean_regional), y = mean_regional)) +
    geom_col(fill = orange) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x = NULL, y = "probability of a child dying before\nreaching age 5, per 1,000 live births") +
    theme_minimal_hgrid()
```

# Second Optional Protocol

Data downloaded from <https://indicators.ohchr.org/>

```{r op2}
op2 <- read_excel(paste0(raw, "/op2.xls"), skip = 1)

pal <- c("Ratified" = "#faa31b", "Not ratified" = "#cccccc")


# op2_export <- op2 %>%
# mutate(rat = ifelse(!is.na(`Date of Ratification/Accession`), T, F),
#        country_cleaned = case_when(Country == "Netherlands" ~ "Netherlands (Kingdom of the)",
#                                    Country == "Côte d'Ivoire" ~ "Côte d’Ivoire",
#                                    Country == "Czech Republic" ~ "Czechia",
#                                   .default = Country)
#       ) %>%
#     left_join(unregions, by = c("country_cleaned" = "Country or Area")) %>%
#     filter(!is.na(`M49 Code`)) %>%
# group_by(`Region Name`) %>%
# summarize(Ratified = sum(rat)/n(),
#           `Not ratified` = 1 - Ratified) %>%
# pivot_longer(cols = c(Ratified, `Not ratified`))

# op2_export %>%
# write.csv(paste0(path, "/op2_pct.csv"), row.names = FALSE)

# op2_export %>%
#   ggplot(aes(x = fct_reorder(`Region Name`, value * (name == "Ratified")), y = value)) +
#   geom_col(aes(fill = name)) +
# scale_fill_manual(values = pal) +
# labs(fill = "Ratified OP2", y = "% of countries in region", x = NULL) +
# scale_y_continuous(expand = c(0, 0)) +
# theme_minimal_hgrid()

# OR

op2_export2 <- op2 %>%
    mutate(
        rat = ifelse(!is.na(`Date of Ratification/Accession`), T, F),
        country_cleaned = case_when(Country == "Netherlands" ~ "Netherlands (Kingdom of the)",
            Country == "Côte d'Ivoire" ~ "Côte d’Ivoire",
            Country == "Czech Republic" ~ "Czechia",
            .default = Country
        )
    ) %>%
    left_join(unregions, by = c("country_cleaned" = "Country or Area")) %>%
    filter(!is.na(`M49 Code`)) %>%
    group_by(`Region Name`) %>%
    summarize(
        Ratified = sum(rat),
        `Not ratified` = sum(!rat)
    ) %>%
    pivot_longer(cols = c(Ratified, `Not ratified`)) %>%
    rename(region = `Region Name`) %>%
    group_by(region) %>%
    mutate(pct_rat = value[name == "Ratified"] / sum(value)) %>%
    ungroup() %>%
    mutate(
        region = fct_reorder(region, pct_rat),
        name = fct_rev(name)
    )

op2_export2 %>%
    write.csv(paste0(path, "/op2_mosaic.csv"), row.names = FALSE)

op2_export2 %>%
    ggplot() +
    geom_mosaic(aes(x = product(name, region), fill = name, weight = value)) +
    scale_fill_manual(values = pal, guide = "none") +
    theme_void() +
    theme(axis.text = element_text()) +
    coord_flip() +
    scale_x_productlist(expand = c(0, 0)) +
    scale_y_productlist(expand = c(0, 0))
```