---
title: "Right to freedom of expression"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
execute:
  warning: false
  message: false
---

```{r setup}
#| echo: false
library(ghRo)
library(tidyverse)
library(here)
library(ggforce)
library(scales)
library(cowplot)
library(ggtext)

raw <- here("data/raw/")

dir.create(here("drafts/rightToFreedomOfExpression/datasets"), recursive = TRUE, showWarnings = FALSE)

out <- here("drafts/rightToFreedomOfExpression/datasets/")
```

Indicators:

- public access to information (SG_INF_ACCSS)

- HRDs

- Internet shutdown (Tier III)

# Access to information

```{r}
eda(getHRDx("SG_INF_ACCSS"))
```

HRDx only has data for 2024. On their site, they interpret this 139 as "139 UN Member States have adopted constitutional, statutory and/or policy guarantees for public access to information."

In the below linked report, for 2025 they say this is "141 UN Member States that have adopted statutory guarantees for public access to information", so this is more updated. 

So 141 of 193 member states - maybe can do a half donut. 

```{r}
data.frame(count = c(141, 193-141), cat = c("yes", "no")) |>
mutate(share = count/sum(count)) |>
mutate(name = factor(cat), 
ymax = cumsum(share), 
ymin = c(0, head(ymax, n= -1))) |>
  mutate_at(vars(starts_with("y")), rescale, to=pi*c(-.5,.5), from=0:1) %>%
  ggplot + 
  geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = .5, r = 1, start = ymin, end = ymax, fill=name)) + 
  scale_fill_manual(values = c("yes" = "#006fb7", "no" = "#cccccc"), guide = "none") +
ggtitle(paste0("In 2024, <b style = 'color:#006fb7;'> 141 countries </b> have adopted constitutional, statutory and/or policy guarantees for public access to information.")) +
  coord_fixed() + 
    theme_nothing() +
    theme(legend.position = "bottom", 
    plot.title = element_textbox_simple(padding = margin(5.5, 5.5, 5.5, 5.5),
    margin = margin(0, 0, 10, 0),
    lineheight = 1, 
    size = 28
))
```

[See the 2025 UNESCO report](https://unesdoc.unesco.org/in/documentViewer.xhtml?v=2.1.196&id=p::usmarcdef_0000398030&highlight=public%20access%20to%20information&file=/in/rest/annotationSVC/DownloadWatermarkedAttachment/attach_import_c6ad0f0c-5a28-4a66-a11d-ab0a4aa3b6e0%3F_%3D398030eng.pdf&locale=en&multi=true&ark=/ark:/48223/pf0000398030/PDF/398030eng.pdf#%5B%7B%22num%22%3A18%2C%22gen%22%3A0%7D%2C%7B%22name%22%3A%22XYZ%22%7D%2Cnull%2Cnull%2Cnull%5D)

```{r}
ati <- data.frame(response = c("Yes", "In progress", "No"), value = c(115, 4, 4))

ati |>
mutate(response = factor(response, levels = c("Yes", "In progress", "No"))) |>
arrange(desc(response)) |>
mutate( 
half = value/2, 
csum = cumsum(lag(value, default = 0)), 
pos = ifelse(response == "No", 1, csum + half)) |>
ggplot(aes(y = "2024", x = value, fill = response, color = response)) + 
geom_col(aes(width = 1)) +
scale_color_manual(values = c("No" = "#B5515B", "In progress" = "#C5C8D4", "Yes" = "#4D7799")) + 
scale_fill_manual(values = c("No" = "#B5515B", "In progress" = "#C5C8D4", "Yes" = "#4D7799")) + 
geom_text(aes(x = pos, label = response), hjust = 0, vjust = 1, nudge_y = -.55, size = 6) + 
geom_text(aes(x = pos, label = value), , color = "white", hjust = 0, vjust = .5, size = 6) + 
scale_y_discrete(expand = expansion(add = 5)) +
scale_x_continuous(expand = c(0, 0)) +
theme_nothing()
```

The actual indicator is a survey of 8 questions, each scored 0 to 2, allowing a total score from 0 to 9. 

26 of 123 respondents scored below 4.5. Average score was 6.32/9. 

Each country's score is available in PDF format so we'd have to download each country individually to obtain analyses further than what is in their 2025 report. 

# killings, disappearances, etc. of HRDs

Will need links to right to life and to right to liberty here, which mention killings and disappearances specifically

Killings = VC_HRD_KIMICRO

Attacks (has disagg) = VC_HRD_KIDISDE

With disappearances, we highlighted the time trend and regional differences

With killings, time trends for world and region

What to do differently here? 

- by sex

- by attack type

So more sums of countries/comparisons as such

```{r}
attacks <- getHRDx("VC_HRD_KIDISDE", 
params = list(where = list(sexCode = "_T", 
datetime = select_years(2024), 
areaType = "Country"))) |> count(value, attackTypeCode) |> 
left_join(hrdxDisaggregations |> filter(disagg == "attackType") |> select(name, code), by = c("attackTypeCode" = "code")) |> 
group_by(name) |> 
mutate(num_w_data = sum(n)) |>
ungroup() |> 
mutate(at_least_1 = case_when(value > 0 ~ "At least 1 attack", 
.default = "Zero attacks")) |> 
select(-value, -attackTypeCode) |>
group_by(at_least_1, name) |>
mutate(at_least_1_n = sum(n), 
share = at_least_1_n/num_w_data) |> 
ungroup() |>
distinct(share, num_w_data, name, at_least_1, at_least_1_n)
```

```{r}
#| fig-width: 8
library(patchwork)
library(ggtext)

plots = list()

for (attack in unique(attacks$name)) {
    num_attacks <- attacks |> filter(name == attack, at_least_1 == "At least 1 attack") |> pull(at_least_1_n)
    print(paste0("There were ", attacks |> filter(name == attack) |> pull(at_least_1_n) |> sum(), " countries with data, and ", num_attacks, " with attacks"))

plots[[attack]] <- attacks |>
filter(name == attack) |>
mutate(name = factor(at_least_1), 
ymax = cumsum(share), 
ymin = c(0, head(ymax, n= -1))) |>
  mutate_at(vars(starts_with("y")), rescale, to=pi*c(-.5,.5), from=0:1) %>%
  ggplot + 
  geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = .5, r = 1, start = ymin, end = ymax, fill=at_least_1)) + 
  coord_fixed() + 
  scale_fill_manual(NULL, values = c("Zero attacks" = "#cccccc", "At least 1 attack" = "#DF2626")) +
  ggtitle(paste0("In 2024, <b style = 'color:#DF2626;'>", num_attacks[1], " countries </b> had at least 1 verified case of ", str_to_lower(attack), " of human rights defenders.")) +
    theme_nothing() +
    theme(legend.position = "bottom", 
    plot.title = element_textbox_simple(padding = margin(5.5, 5.5, 5.5, 5.5),
    margin = margin(0, 0, 10, 0),
    lineheight = 1, 
    size = 14
))

}

wrap_plots(plots) + 
plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

```{r}
library(jsonlite)
library(httr)

get_unsdg <- function(code) {
res <- GET(url = paste0("https://unstats.un.org/sdgs/UNSDGAPIV5/v1/sdg/Series/Data?seriesCode=", code))

n_pages <- fromJSON(content(res, "text"), flatten = TRUE)$totalPages

paged_data <- list()

for (n in 1:n_pages) {
  page_res <- GET(url = paste0("https://unstats.un.org/sdgs/UNSDGAPIV5/v1/sdg/Series/Data?seriesCode=", code, "&page=", n))
  d <- fromJSON(content(page_res, "text"), flatten = TRUE)$data
  paged_data[[n]] <- d
}

rbind_pages(paged_data) |> mutate(value = as.integer(value))
}

enf_dis <- get_unsdg("VC_VOC_ENFDIS")

killings <- get_unsdg("VC_VAW_MTUHRA")
```

```{r}
#| fig-height: 10
library(waffle)
library(showtext)

world_ed <- enf_dis |> filter(geoAreaName == "World") |> select(value, dimensions.Sex, geoAreaCode, timePeriodStart) |> mutate(attackType = "Enforced disappearance")
world_k <- killings |> filter(geoAreaName == "World") |> select(value, dimensions.Sex, geoAreaCode, timePeriodStart) |> mutate(attackType = "Killing")

font_add(
    family = "FontAwesome5Free-Solid",
    regular = system.file("fonts", "fa-solid-900.ttf", package = "waffle")
)
showtext_auto()

pal <- c("Killing" = ghro_palettes$cats2[1], "Enforced disappearance" = ghro_palettes$cats2[2])

rbind(world_ed, world_k) |> 
filter(dimensions.Sex == "BOTHSEX", timePeriodStart %in% c(2019, 2024)) |>
    ggplot(aes(label = attackType, values = value)) +
    geom_pictogram(
        aes(colour = attackType),
        family = "FontAwesome5Free-Solid",
        n_rows = 10,
        size = 3, 
        flip = TRUE
    ) +
    scale_label_pictogram(
        name = NULL,
        values = c(
           "user"        ), 
    ) +
    scale_color_manual(NULL, values = pal) +
    facet_wrap(~timePeriodStart) + 
    coord_equal() +
    theme_void() +
    theme(legend.position = "top", legend.direction = "vertical")
```

For d3:
```{r}
rbind(world_ed, world_k) |> 
filter(dimensions.Sex == "BOTHSEX", timePeriodStart %in% c(2019, 2024)) |> 
mutate(cat = case_when(str_detect(attackType, "Kill") ~ "kil", 
.default = "edis")) |>
select(timePeriodStart, cat, value) |>
rename(year = timePeriodStart) |>
pivot_wider(names_from = cat, values_from = value) |>
write.csv(paste0(out, "/d3_pictogram_FOE.csv"), row.names = FALSE)
```
I think it might be really interesting/hit harder to have a visualization that keeps going to show the scale

```{r}
rbind(world_ed, world_k) |> 
filter(dimensions.Sex == "BOTHSEX", timePeriodStart %in% c(2019, 2024), attackType == "Killing") |>
    ggplot(aes(label = attackType, values = value)) +
    geom_pictogram(
        aes(colour = attackType),
        family = "FontAwesome5Free-Solid",
        n_rows = 10,
        size = 3, 
        flip = TRUE
    ) +
    scale_label_pictogram(
        name = NULL,
        values = c(
           "user"        ), 
    ) +
    scale_color_manual(NULL, values = pal) +
    facet_wrap(~timePeriodStart) + 
    coord_equal() +
    theme_void() +
    theme(legend.position = "top", legend.direction = "vertical")
```

~ scrollytelling ~

```{r}
rbind(world_ed, world_k) |> 
filter(dimensions.Sex != "BOTHSEX", timePeriodStart != 2025) |>
group_by(timePeriodStart, attackType) |>
mutate(pct = value/sum(value)*100) |> 
select(dimensions.Sex, timePeriodStart, attackType, pct) |>
pivot_wider(names_from = "dimensions.Sex", values_from = pct)
```

Slight increase in % of defenders that are women from 2023 to 2024 - but very slight. 
Might be better to just say that 2 of 7 hrds killed/disappeared is a woman

When we have disaggregated data, can also explore that here. 

Since there's so much in right to life and plenty in right to liberty, going to put regional charts here. Global trends can stay with those respective articles

```{r}
enf_dis_regions <-  enf_dis |> filter(geoAreaName != "World", timePeriodStart == 2024, dimensions.Sex == "BOTHSEX") |>  mutate(value = as.integer(value)) |> filter( !is.na(value))

enf_dis_regions |> 
ggplot(aes(x = value, y = forcats::fct_reorder(stringr::str_wrap(geoAreaName, 20), value))) + 
geom_col(fill = "#006fb7") + 
theme_ghro_bar("horizontal", font_size = 28, x_label = "Number of enforced disappearances\nof human rights defenders") + 
theme(text = element_text(lineheight = .6))
```

```{r}
killings_regions <-  killings |> filter(geoAreaName != "World", timePeriodStart == 2024, dimensions.Sex == "BOTHSEX") |>  mutate(value = as.integer(value)) |> filter( !is.na(value))

killings_regions |> 
ggplot(aes(x = value, y = forcats::fct_reorder(stringr::str_wrap(geoAreaName, 20), value))) + 
geom_col(fill = "#006fb7") + 
theme_ghro_bar("horizontal", font_size = 28, x_label = "Number of killings\nof human rights defenders") + 
theme(text = element_text(lineheight = .6))
```

```{r regional-hrds-line}
#| fig-height: 7

hrds <- readxl::read_excel(here(raw, "hrd_killings_updated2025.xlsx")) |>
filter(Sex == "BOTHSEX") |>
select(-Goal, -Target, -Indicator, -SeriesCode, -Sex, -Units, -`Reporting Type`) |> 
pivot_longer(cols = `2015`:`2025`, names_to = "year") |>
mutate(value = as.numeric(value))

world_pop <- 8300678396
lac_pop <- 672135562
nawa_pop <- 599872197
cat("Latin America and the Caribbean is", scales::percent(lac_pop/world_pop), "of the world population.")
cat("North Africa and Western Asia is", scales::percent(nawa_pop/world_pop), "of the world population.")

hrds_ts <- hrds |> 
filter(GeoAreaCode != 1, year != 2025)

labels <- hrds_ts |> filter(year == 2024, GeoAreaName != "Oceania")

hrds_ts |>
filter(GeoAreaName != "Oceania") |>
ggplot(aes(x = as.integer(year), y = value, color = GeoAreaName, group = GeoAreaName)) + 
geom_line(linewidth = .75) + 
ggrepel::geom_text_repel(data = labels, aes(label = str_wrap(GeoAreaName, 20)), hjust = 0, size = 8, lineheight = .5, xlim = c(2024, Inf)) +
scale_color_ghro("cats2", guide = "none") +
theme_ghro_line(font_size = 28) +
scale_x_continuous(NULL, breaks = seq(2015, 2024, by = 3)) +
scale_y_continuous("Number of human rights defenders killed") +
theme(plot.margin = margin(r = 125, b = 5)) +
coord_cartesian(clip = "off")
```

```{r}
enf_dis_ts <- enf_dis |> 
filter(geoAreaCode != 1, timePeriodStart > 2019, timePeriodStart != 2025, dimensions.Sex == "BOTHSEX", geoAreaName != "Oceania") |> select(geoAreaName, timePeriodStart, value) |>
drop_na()
labels <- enf_dis_ts |> filter(timePeriodStart == 2024)

enf_dis_ts |>
ggplot(aes(x = as.integer(timePeriodStart), y = value, color = geoAreaName, group = geoAreaName)) + 
geom_line(linewidth = .75) + 
ggrepel::geom_text_repel(data = labels, aes(label = str_wrap(geoAreaName, 20)), hjust = 0, size = 8, lineheight = .5, xlim = c(2024, Inf)) +
scale_color_ghro("cats2", guide = "none") +
theme_ghro_line(font_size = 28) +
scale_x_continuous(NULL) +
scale_y_continuous("Number of human rights defenders disappeared") +
theme(plot.margin = margin(r = 125, b = 5)) +
coord_cartesian(clip = "off")
```