---
title: "Right to a fair trial"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
execute:
  warning: false
  message: false
include-in-header:
  - text: |
      <style>
      .cell-output-stdout code {
        word-break: break-wor !important;
        white-space: pre-wrap !important;
      }
      </style>
---

```{r setup}
#| echo: false
library(ghRo)
library(tidyverse)
library(here)
library(ggforce)
library(scales)
library(cowplot)
library(ggtext)

raw <- here("data/raw/")

dir.create(here("drafts/rightToFairTrial/datasets"), recursive = TRUE, showWarnings = FALSE)

out <- here("drafts/rightToFairTrial/datasets/")
```

```{r}
regionIds <- regions |> filter(level %in% 0:1)
unsentenced <- getHRDx("VC_PRS_UNSNT", params = list(where = list(datetime = select_years(years = 2010:2026)), regionId = regionIds))

# eda(unsentenced, detailed = TRUE)
```

```{r}
library(waffle)
library(magrittr)
library(showtext)
font_add(
    family = "FontAwesome5Free-Solid",
    regular = system.file("fonts", "fa-solid-900.ttf", package = "waffle")
)
showtext_auto()

unsent <- data.frame(name = c("sentenced", "unsentenced"), value = c(7, 3))

unsent |> 
    ggplot(aes(label = name, values = value)) +
    geom_pictogram(
        aes(colour = name),
        family = "FontAwesome5Free-Solid",
        n_rows = 1,
        size = 35
    ) +
    scale_label_pictogram(
        name = NULL,
        values = c(
            "sentenced" = "user",
            "unsentenced" = "user"
        )
    ) +
    scale_color_manual(
        values = c(
            "unsentenced" = "#006fb7",
            "sentenced" = "#f5f5f1"
        )
    ) +
    coord_equal(clip = "off") +
    theme_void() +
    theme(legend.position = "none") +
    scale_y_continuous(limits = c(.9, 1.5))
```


1. % of prison population that is unsentenced has been increasing since 2010, was decreasing between 2016 and 2019, but then spiked to an all time high in 2021 of 31.7%, meaning that more than 3 in 10 inmates globally were unsentenced. Since 2021 it has declined, but remains above 30%.

```{r}
ghro_line("VC_PRS_UNSNT", level = "regional", line_args = list(linewidth = .75), theme_args = list(font_size = 28), region_label_size = 9) + theme(plot.margin = margin(r = 50))
```

2. Increase was most obvious in Oceania, highest in the world now

```{r}
ghro_region_bar("VC_PRS_UNSNT", theme_args = list(font_size = 28))
```

1. Nearly 40% in Oceania

Gendered differences? 
```{r}
unsentenced |>
filter(sexCode != "_T", datetime == max(datetime)) |> 
inner_join(regions |> filter(level == 1), by = "regionId") |>
ggplot(aes(x = value, y = fct_reorder(name, value))) + 
geom_line(aes(group = name), linewidth = .75) +
geom_point(aes(color = sexCode), size = 5) + 
scale_color_ghro("gender", name = NULL, labels = c("F" = "Female", "M" = "Male")) + 
theme_ghro_bar(orientation = "horizontal", font_size = 28) +
theme(legend.position = "top") +
scale_x_continuous("% of prison population that is unsentenced", labels = scales::percent_format(scale = 1))
```

1. Gender gap is large in Oceania and Africa - in both regions, women are about 1/4 more likely to be unsentenced than men