---
title: "Right to non-discrimination"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
execute:
  warning: false
  message: false
  out-width: 100%
---

```{r setup}
#| echo: false
library(waffle)
library(magrittr)
library(showtext)
library(hrbrthemes)
library(tidyverse)
library(waffle)
library(tidyverse)
library(cowplot)
library(rvest)
library(ggbeeswarm)
library(scales)
library(ggtext)
library(httr)
library(jsonlite)
library(ggrepel)
library(readxl)
library(ggmosaic)
library(here)

dir.create(here("drafts/rightToNondiscrimination/datasets"), recursive = TRUE, showWarnings = FALSE)
path <- here("drafts/rightToNondiscrimination/datasets/")
raw <- here("data/raw/")

perceptions <- read.csv(paste0(raw, "/expofdisc2026.csv")) %>%
    select(-starts_with("X"))
```

```{r}
font_add(
    family = "FontAwesome5Free-Solid",
    regular = system.file("fonts", "fa-solid-900.ttf", package = "waffle")
)
showtext_auto()

perceptions %>%
    filter(Type == "Global") %>%
    mutate(experienced.discrimination = round(Value)) %>%
    select(experienced.discrimination) %>%
    mutate(didnt.experience.discrimination = 100 - experienced.discrimination) %>%
    pivot_longer(everything()) %>%
    mutate(value = value / 20) %>%
    ggplot(aes(label = name, values = value)) +
    geom_pictogram(
        aes(colour = name),
        family = "FontAwesome5Free-Solid",
        n_rows = 1,
        size = 35
    ) +
    scale_label_pictogram(
        name = NULL,
        values = c(
            "experienced.discrimination" = "user",
            "didnt.experience.discrimination" = "user"
        )
    ) +
    scale_color_manual(
        values = c(
            "experienced.discrimination" = "#faa31b",
            "didnt.experience.discrimination" = "#f5f5f1"
        )
    ) +
    coord_equal() +
    theme_void() +
    theme(legend.position = "none") +
    scale_y_continuous(limits = c(.75, 2))
```

```{r}
library(ghRo)

gender_ts <- perceptions |> filter(Sex != "BOTHSEX", Grounds.of.discrimination == "ALL", Age == "ALLAGE", Disability.status == "_T", Location == "ALLAREA", Education == "_T", Income == "_T", TimePeriod > 2015) |> 
select(TimePeriod, Sex, GeoAreaCode, Value) |>
pivot_wider(names_from = "Sex", values_from = "Value") |>
# first, we are dropping any country-year pairs that don't have data for both men and women
drop_na() |>
# then, we are binning years
mutate(bin = case_when(between(TimePeriod, 2016, 2020) ~ "2016-2020", 
between(TimePeriod, 2021, 2025) ~ "2021-2025", 
.default = NA)) |>
group_by(GeoAreaCode, bin) |>
# for each country and time range, only take most most recent
filter(TimePeriod == max(TimePeriod)) |>
ungroup() |> 
pivot_longer(c(FEMALE, MALE), names_to = "Sex") |>
select(-TimePeriod) |> 
# finally, we only take countries that have data for both time bins
pivot_wider(names_from = "bin", values_from = "value") |>
drop_na() |>
pivot_longer(c(`2016-2020`, `2021-2025`)) |>
group_by(Sex, name) |>
summarize(m = mean(value)) |>
mutate(Sex = str_to_title(Sex)) |>
pivot_wider(names_from = "name", values_from = "m") 

gender_ts |> 
write.csv(paste0(path, "/time_change_sex_discrimination.csv"), row.names = FALSE)

gender_ts |>
pivot_longer(c(`2016-2020`, `2021-2025`)) |>
mutate(Sex = ifelse(Sex == "Female", "F", "M")) |>
ungroup() |>
ggplot(aes(x = name, y = value, group = Sex, color = Sex)) + 
geom_line() + 
geom_point() + 
scale_color_ghro("gender") + 
theme_ghro_bar()
```

Methodology was as follows:
1. For each country, took the most recent data value from 2016-2020 and then for 2021-2025. For example, if Argentina had values for 2017 and 2019, its value for 2016-2020 is 2019. 
2. Then, I narrowed it down to only countries that had data for both time periods.
There were 27 countries that had data a) for both male and female for the same year, and b) for both time periods. 


```{r}
discrim_pd <- data.frame(name = c("experienced.discrimination", "didnt.experience.discrimination"), value = c(3, 7))

discrim_pd |> 
    ggplot(aes(label = name, values = value)) +
    geom_pictogram(
        aes(colour = name),
        family = "FontAwesome5Free-Solid",
        n_rows = 1,
        size = 30
    ) +
    scale_label_pictogram(
        name = NULL,
        values = c(
            "experienced.discrimination" = "wheelchair",
            "didnt.experience.discrimination" = "wheelchair"
        )
    ) +
    scale_color_manual(
        values = c(
            "experienced.discrimination" = "#faa31b",
            "didnt.experience.discrimination" = "#f5f5f1"
        )
    ) +
    coord_equal() +
    theme_void() +
    theme(legend.position = "none") +
    scale_y_continuous(limits = c(.9, 1.75))
```

```{r}
# data directly from 16.b.1 underlying data - 2026 spreadsheet, tab "5 - gender"
sex_grounds <- data.frame(
  Ground = c(
    "Age",
    "Disability or health status",
    "Ethnicity, colour, language",
    "Geographic location or place of residence",
    "Marital and family status",
    "Migration status",
    "Other grounds",
    "Political opinion",
    "Religion",
    "Sex",
    "Sexual orientation and gender identity",
    "Socio-economic status"
  ),
  FEMALE = c(4.52, 1.78, 4.83, 3.66, 5.21, 3.89, 4.14, 2.29, 4.13, 5.55, 2.15, 4.79),
  MALE = c(4.38, 2.05, 5.52, 3.80, 2.84, 3.18, 4.14, 3.09, 4.64, 2.45, 2.01, 4.76)
) |> 
mutate(diff = FEMALE - MALE) 

export_sex_grounds <- sex_grounds

colnames(export_sex_grounds) <- str_to_title(colnames(export_sex_grounds))

export_sex_grounds|> write.csv(paste0(path, "/sex_discrim_grounds.csv"), row.names = FALSE)

sex_grounds |>
pivot_longer(c(FEMALE, MALE)) |>
mutate(ground = str_wrap(Ground, 25)) |>
ggplot(aes(x = value, y = fct_reorder(ground, diff), color = name)) + 
geom_line(aes(group = ground), color = "grey70", linewidth = 0.8) +

geom_point(size = 5) + 
scale_color_manual(values = c("FEMALE" = ghro_palettes$gender[["F"]], "MALE" = ghro_palettes$gender[["M"]]), guide = "none") +
labs(
  title = "Perceived discrimination by ground",
  subtitle = glue::glue(
    "Sorted from grounds where <span style='color:{ghro_palettes$gender[['F']]}'>**women**</span> report the most discrimination relative to men, to those where <span style='color:{ghro_palettes$gender[['M']]}'>**men**</span> report more"
  ),
  x = "Prevalence of discrimination", y = NULL
) +
scale_x_continuous(labels = scales::percent_format(scale = 1)) +
theme_minimal_hgrid() +
theme(plot.subtitle = element_markdown()) +
annotate("text", x = -Inf, y = Inf, label = "↑ larger female gap", 
         hjust = -0.1, vjust = 1.5, size = 3, fontface = "italic", color = "grey40") +
annotate("text", x = -Inf, y = -Inf, label = "↓ larger male gap", 
         hjust = -0.1, vjust = -0.5, size = 3, fontface = "italic", color = "grey40")
```


# Gender equality frameworks

```{r}
urls <- c("lfp", "emp", "vaw", "mar")
labels <- c("overarching", "employment", "violence", "marriage")
params <- list(
    region = TRUE,
    country = TRUE,
    year = TRUE,
    series = TRUE,
    natureEstimate = TRUE
)

gender_eq <- list()

for (i in 1:length(urls)) {
    body <- list(
        limit = 1000
    )

    response <- POST(
        url = paste0("https://hrdx-api-staging.un.org/sgLglGeneq", urls[i], "/search"),
        body = toJSON(body, auto_unbox = TRUE),
        content_type_json()
    )

    results <- fromJSON(content(response, "text"), flatten = TRUE)$results

    results$name <- labels[[i]]

    gender_eq[[labels[i]]] <- results
}
```

gender equality frameworks

```{r}
gender_equality <- gender_eq %>%
    bind_rows() %>%
    mutate(name = case_when(
        name == "overarching" ~ "overarching legal frameworks and public life",
        name == "employment" ~ "employment and economic benefits",
        name == "violence" ~ "violence against women",
        name == "marriage" ~ "marriage and family"
    ), 
    year = year(datetime))
```

frameworks over time - global

frameworks 2024 regions
```{r}
ge_regions <- gender_equality |> 
filter(year == 2024, !is.na(countryId)) |> 
left_join(ghRo::countries |> rename(country = name), by = c("countryId" = "countryId")) |>
select(name, year, level1, value) |>
mutate(level1 = as.character(level1)) |>
left_join(ghRo::regions |> select(regionId, name) |> rename(region = name), by = c("level1" = "regionId")) |>
group_by(region, name) |>
summarize(avg_unwt = mean(value), 
n = n()) 

ge_regions |> select(-n) |> pivot_wider(names_from = region, values_from = avg_unwt) |> 
select(all_of(c("name","Europe","Africa", "Asia","Oceania", "Americas"))) |> 
mutate(name = factor(name, levels = c("overarching legal frameworks and public life", "violence against women", "employment and economic benefits", "marriage and family"))) |> arrange(name) |> write.csv(paste0(path, "/gender_equality_regions.csv"), row.names = FALSE)

ge_regions |>
ggplot(aes(x = avg_unwt, y = fct_reorder(region, avg_unwt))) + 
geom_col(aes(fill = region)) + 
facet_wrap(~name, ncol = 1) + 
scale_fill_ghro("regions") + 
guides(fill = "none") +
scale_x_continuous("average regional achievement of legal frameworks, 2024", expand = c(0, 0)) +
scale_y_discrete(NULL) +
theme_minimal_vgrid() + 
theme(axis.ticks = element_blank())
```
