---
title: "Right to education"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
execute:
  warning: false
  message: false
  out-width: 100%
---

```{r setup}
#| echo: false
library(magrittr)
library(showtext)
library(hrbrthemes)
library(waffle)
library(tidyverse)
library(cowplot)
library(rvest)
library(ggbeeswarm)
library(scales)
library(httr)
library(jsonlite)
library(ggrepel)
library(readxl)
library(ggmosaic)
library(ggarrow)
library(here)
library(ggtext)
source(here("consts.R"))

dir.create(here("drafts/rightToEducation/datasets"), recursive = TRUE, showWarnings = FALSE)
path <- here("drafts/rightToEducation/datasets/")
raw <- here("data/raw/")
raw_right <- here("data/raw/rightToEducation/")
```

# Indicators

- Functional skills

  - Population achieving minimum functional skills proficiency (%) (SE_ADT_FUNS)

  - Immigration status parity in functional skills (ratio) (SE_IMP_FPOF)

  - Gender parity in functional skills (ratio) (SE_TOT_GPI_FS)

  - Socio-economic parity in functional skills (ratio) (SE_TOT_SESPI_FS)

- Completion rate

  - Education completion rate (%) (SE_TOT_CPLR)

  - Gender parity in completion rate (ratio) (SE_AGP_CPRA)

  - Location parity in education completion rate (ratio) (SE_ALP_CPLR)

  - Wealth parity in education completion rate (ratio) (SE_AWP_CPRA)

- ICT skills

  - Gender parity in ICT skills (ratio) (SE_GPI_ICTS)

- Education and training

  - Gender parity in education and training participation (ratio) (SE_GPI_PART)

- Organized learning

  - Gender parity in organized learning participation (ratio) (SE_GPI_PTNPRE)

- Qualified teachers

  - Teachers with minimum required qualifications (%) (SE_TRA_GRDL)

  - Gender parity in qualified teachers (ratio) (SE_GPI_TCAQ)

- Reading and mathematics

  - Children and youth achieving minimum reading and math proficiency (%) (SE_TOT_PRFL)

  - Gender parity in reading and mathematics (ratio) (SE_TOT_GPI)

  - Language test parity in reading and mathematics (ratio) (SE_LGP_ACHI)

  - Immigration status parity in reading and mathematics (ratio) (SE_NAP_ACHI)

  - Rural-urban parity in reading and mathematics (ratio) (SE_TOT_RUPI)

  - Socio-economic parity in reading and mathematics (ratio) (SE_TOT_SESPI)

- Youth not in education

  - Youth not in education, employment or training, 19th ICLS (%) (SL_TLF_NEET_19ICLS)

```{r}
#| include: false
#| echo: false
# getting all the data from HRDx

right_codes <- c(
  "SE_ADT_FUNS",
  "SE_AGP_CPRA",
  "SE_ALP_CPLR",
  "SE_AWP_CPRA",
  "SE_GPI_ICTS",
  "SE_GPI_PART",
  "SE_GPI_PTNPRE",
  "SE_GPI_TCAQ",
  "SE_TOT_GPI",
  "SE_IMP_FPOF",
  "SE_LGP_ACHI",
  "SE_NAP_ACHI",
  "SE_TOT_CPLR",
  "SE_TOT_GPI_FS",
  "SE_TOT_PRFL",
  "SE_TOT_RUPI",
  "SE_TOT_SESPI",
  "SE_TOT_SESPI_FS",
  "SE_TRA_GRDL",
  "SL_TLF_NEET_19ICLS"
)

right_raw_data <- list()
for (d in right_codes) {
    right_raw_data[[d]] <- getHRDx(d, offline = TRUE, raw = raw_right)
}
```

# Completion rate

```{r}
#| include: false
#| echo: false
comp_rt <- right_raw_data[['SE_TOT_CPLR']] |>
mutate(year = year(datetime))

dim(comp_rt)
colnames(comp_rt) # disaggs available are sexCode, urban/rural, quantile, education, 

length(unique(comp_rt$countryId)) # we have data for 168 countries

unique(comp_rt$regionId) # and regions, mostly sub-regions

sort(unique(year(comp_rt$datetime))) # data for 2010-2025

for (disagg in c("sexCode", "dwellingUrbanisationCode", 
                "educationCode", "quantileCode")) {
    print(table(comp_rt[[disagg]]))
                }
```

```{r}
comp_world <- comp_rt |>
    filter(regionId == 1) |>
    left_join(edu, by = c("educationCode" = "code"))

pal <- palettes$education(11)

comp_world |>
    filter(sexCode == "_T", dwellingUrbanisationCode == "_T", quantileCode == "_T") |>
    ggplot(aes(x = year, y = value)) +
    geom_line(aes(color = factor(name, levels = c("Primary education", "Lower secondary education", "Upper secondary education")))) +
    scale_color_manual("Education level", values = c("Primary education" = pal[3], "Lower secondary education" = pal[6], "Upper secondary education" = pal[11])) +
    scale_y_continuous("education completion rate", labels = scales::percent_format(scale = 1)) +
    labs(x = NULL) +
    theme_minimal_grid()

comp_world |>
    filter(sexCode != "_T", dwellingUrbanisationCode == "_T", quantileCode == "_T", year == max(year)) |>
    ggplot(aes(x = factor(name, levels = c("Primary education", "Lower secondary education", "Upper secondary education"), labels = c("Primary education", "Lower secondary\neducation", "Upper secondary\neducation")), y = value)) +
    geom_col(aes(fill = sexCode), position = "dodge") +
    scale_fill_ghro("gender") +
    scale_y_continuous("education completion rate", labels = scales::percent_format(scale = 1), expand = c(0, 0)) +
    labs(x = NULL) +
    theme_minimal_hgrid()
```

1. Levels of completion drop as education level advances

2. Girls tend to have higher levels of completion across all levels monitored. 

```{r}
#| include: false
#| echo: false
comp_loc <- right_raw_data[["SE_ALP_CPLR"]] |>
    mutate(year = year(datetime))

comp_gen <- right_raw_data[["SE_AGP_CPRA"]] |>
    mutate(year = year(datetime))

    comp_wea <- right_raw_data[["SE_AWP_CPRA"]] |>
    mutate(year = year(datetime))


dim(comp_loc)
colnames(comp_loc) # disaggs available are sexCode, quantile, education
colnames(comp_gen) # disaggs available are dwellingUrbanisationCode, quantile, education,
colnames(comp_wea)

length(unique(comp_loc$countryId)) # we have data for 168 countries

unique(comp_loc$regionId) # and no regions

sort(unique(year(comp_loc$datetime))) # data for 2010-2024
```

```{r}
comp_loc_world <- comp_loc |>
    filter(sexCode == "_T", quantileCode == "_T", year == max(year)) |>
    group_by(year, educationCode) |>
    summarize(m = symmetrize_ratio(mean(value)), 
    n = n()) |>
    left_join(edu, by = c("educationCode" = "code")) |>
    ungroup() |>
    mutate(name = factor(name, levels = c("Primary education", "Lower secondary education", "Upper secondary education"), labels = c("Primary education", "Lower secondary\neducation", "Upper secondary\neducation")))

comp_loc_world |>
    ggplot(aes(x = m, y = name)) +
                     geom_vline(xintercept = 1, color = "gray30", linetype = "dashed") +
    geom_segment(aes(x = m, y = name, yend = name), xend = 1, color = "#7e9e9d") + 
    geom_point(color = "#7e9e9d") + 
    scale_x_continuous(NULL, limits = c(0, 2), expand = c(0, 0),  labels = function(x) {
        ifelse(
            x == .5,
            "Higher\n← completion rate\nin urban areas",
            ifelse(x == 1.5, "Higher\ncompletion rate →\nin rural areas", ifelse(x == 1, "Parity", x))
        )
    }) + 
    scale_y_discrete(NULL) +
    theme_minimal_vgrid()
```

1. As education level advances, the difference in rate of completion between urban and rural students grows/becomes more disparate. 

2. Possibility of a trend where the pandemic resulted in the decline, particularly in upper secondary?

```{r}
comp_gen_world <- comp_gen |>
    filter(dwellingUrbanisationCode == "_T", quantileCode == "_T", year == max(year)) |>
    group_by(year, educationCode) |>
    summarize(m = symmetrize_ratio(mean(value))) |>
    left_join(edu, by = c("educationCode" = "code")) |>
    ungroup() |>
    mutate(name = factor(name, levels = c("Primary education", "Lower secondary education", "Upper secondary education"), labels = c("Primary education", "Lower secondary\neducation", "Upper secondary\neducation")))

comp_gen_world |>
     ggplot(aes(x = m, y = name)) +
                      geom_vline(xintercept = 1, color = "gray30", linetype = "dashed") +
    geom_segment(aes(x = m, y = name, yend = name), xend = 1, color = "#990099") + 
    geom_point(color = "#990099") + 
    scale_x_continuous(NULL, limits = c(0, 2), expand = c(0, 0),  labels = function(x) {
        ifelse(
            x == .5,
            "Higher\n← completion rate\nfor boys",
            ifelse(x == 1.5, "Higher\ncompletion rate →\nfor girls", ifelse(x == 1, "Parity", x))
        )
    }) + 
    scale_y_discrete(NULL) +
    theme_minimal_vgrid()
```

1. Globally, girls and boys are about as likely to complete education at all levels, though there is a slight trend towards girls completing upper secondary at a higher rate. 

```{r}
comp_wea_world <- comp_wea |>
    filter(dwellingUrbanisationCode == "_T", sexCode == "_T", year == max(year)) |>
    group_by(year, educationCode) |>
    summarize(m = symmetrize_ratio(mean(value))) |>
    left_join(edu, by = c("educationCode" = "code")) |>
    ungroup() |>
    mutate(name = factor(name, levels = c("Primary education", "Lower secondary education", "Upper secondary education"), labels = c("Primary education", "Lower secondary\neducation", "Upper secondary\neducation")))

colors <- palettes$wealth(5)

comp_wea_world |>
         ggplot(aes(x = m, y = name)) +
                 geom_vline(xintercept = 1, color = "gray30", linetype = "dashed") +

    geom_segment(aes(x = m, y = name, yend = name), xend = 1, color = "#825516") + 
    geom_point(color = "#825516") + 
    scale_x_continuous(NULL, limits = c(0, 2), expand = c(0, 0),  labels = function(x) {
        ifelse(
            x == .5,
            "Higher\n← completion rate\nfor wealthier children",
            ifelse(x == 1.5, "Higher\ncompletion rate →\nfor poorer children", ifelse(x == 1, "Parity", x))
        )
    }) + 
    scale_y_discrete(NULL) +
    theme_minimal_vgrid()
```

1. Gap here is the widest between education levels, and also most disparity between advantaged and disadvantaged groups. 

2. Wealthier children are more than twice as likely to complete upper secondary education as poorer children. 

# Youth not in education

```{r}
#| include: false
#| echo: false
neet <- getHRDx('SL_TLF_NEET', list(where = list(datetime = paste0("01-01-", 2010:2026)))) |>
mutate(year = year(datetime))

neet |> write.csv(here("data/raw/rightToEducation/SL_TLF_NEET.csv"))

dim(neet)
colnames(neet) # disaggs for sex

length(unique(neet$countryId)) # we have data for 182 countries

unique(neet$regionId) # and no regions

sort(unique(year(neet$datetime))) # data for 2010-2025
```

```{r}
world_neet <- neet |>
filter(regionId == 1, sexCode == "_T")

world_neet |>
ggplot(aes(x = year, y = value)) +
geom_line(color = "#006fb7") +
        geom_hline(yintercept = 0, color = "gray30", linetype = "dashed") +
        annotate("text", x = 2025.25, y = 0, hjust = 0, label = "Target") +
                annotate("text", x = 2025.25, y = world_neet |> filter(year == 2025) |> pull(value), hjust = 0, label = world_neet |> filter(year == 2025) |> pull(value) |> scales::percent(scale = 1), color = "#006fb7") +
scale_x_continuous(NULL, expand = c(0, 0)) +
scale_y_continuous("Proportion of youth not in education,\nemployment or training", labels = scales::percent_format(scale = 1), limits = c(0, 50), expand = expansion(add = .25)) +
theme_minimal_grid() +
theme(plot.margin = margin(r = 50, t = 10)) +
coord_cartesian(clip = "off")
```

1. 1 in 5 children globally are not in education, employment or training. 

2. Not much change or improvement in the past 15 years

```{r}
world_sex_neet <- neet |>
filter(regionId == 1, sexCode != "_T")

world_sex_neet |>
ggplot(aes(x = year, y = value)) +
geom_line(aes(color = sexCode)) +
        geom_hline(yintercept = 0, color = "gray30", linetype = "dashed") +
        annotate("text", x = 2025.25, y = 0, hjust = 0, label = "Target") +
geom_text(data = world_sex_neet |> filter(year == max(year)), aes(color = sexCode, label = paste0(ifelse(sexCode == "F", "Female", "Male"), "\n", scales::percent(value, scale = 1))), x = 2025.25, hjust = 0) +
scale_color_ghro("gender") +
scale_x_continuous(NULL, expand = c(0, 0)) +
scale_y_continuous("Proportion of youth not in education,\nemployment or training", labels = scales::percent_format(scale = 1), limits = c(0, 50), expand = expansion(add = .25)) +
guides(color = "none") +
theme_minimal_grid() +
theme(plot.margin = margin(r = 50, t = 10)) +
coord_cartesian(clip = "off")
```

1. Disaggregation by sex reveals that this number rises to more than 1 in 4 for girls, and drops to over 1 in 10 for boys. 

```{r}
region_neet <- neet |> 
filter(year == max(year), sexCode == "_T") |> 
mutate(regionId = as.numeric(regionId)) |> 
inner_join(regions, by = "regionId")

world_val <- region_neet |> 
filter(regionId == 1) |>
pull(value)

region_neet |>
filter(regionId != 1) |>
ggplot(aes(x = fct_reorder(name, value), y = value)) + 
geom_col(fill = "#006fb7") +
        geom_hline(yintercept = world_val, color = "gray30", linetype = "dashed") +
        annotate("text", x = 2, y = world_val + .1, vjust = 0, label = paste0("Global average: ", scales::percent(world_val, scale = 1))) +
scale_y_continuous("Proportion of youth not in education,\nemployment or training", expand = c(0, 0), labels = scales::percent_format(scale = 1)) +
scale_x_discrete(NULL) +
theme_minimal_hgrid() 
```

1. Europe has the lowest proportion (best), whereas the Asia and Africa are higher than the global average and are the most "behind"

# Skills

## ICT skills

```{r}
#| include: false
#| echo: false
gender_ict <- right_raw_data[['SE_GPI_ICTS']] |> 
mutate(year = year(datetime))

dim(gender_ict)
colnames(gender_ict) # skillTypeCode
length(unique(gender_ict$countryId)) # we have data for 102 countries

unique(gender_ict$regionId) # and no regions

sort(unique(year(gender_ict$datetime))) # data for 2010-2025

basic <- gender_ict |> 
filter(skillTypeCode %in% c("SKILL_ICTINFOBAS", "SKILL_ICTCOMMBAS", "SKILL_ICTSAFEBAS", "SKILL_ICTPROBBAS", "SKILL_ICTCONTBAS"), 
year == 2024) |> 
left_join(countryISO, by = c("countryId" = "Code")) |>
left_join(getHRDx("skillType"), by = c("skillTypeCode" = "code")) |>
group_by(name) |>
summarize(m = symmetrize_ratio(mean(value))) |>
mutate(name = trimws(str_remove_all(name, "ICT Skill: Basic|skills")), 
col = case_when(between(m, 1 - .01, 1.01) ~ "P", 
m < 1 - .01 ~ "M", 
.default = "F"))

basic <- gender_ict |> 
filter(skillTypeCode %in% c("SKILL_ICTINFOBAS", "SKILL_ICTCOMMBAS", "SKILL_ICTSAFEBAS", "SKILL_ICTPROBBAS", "SKILL_ICTCONTBAS"), 
year == 2024) |> 
left_join(countryISO, by = c("countryId" = "Code")) |>
left_join(getHRDx("skillType"), by = c("skillTypeCode" = "code")) |>
group_by(name) |>
summarize(m = symmetrize_ratio(mean(value))) |>
mutate(name = trimws(str_remove_all(name, "ICT Skill: Basic|skills")), 
col = ifelse(m > 1, "F", "M"))

above_basic <- gender_ict |> 
filter(skillTypeCode %in% c("SKILL_ICTINFOABAS", "SKILL_ICTCOMMABAS", "SKILL_ICTSAFEABAS", "SKILL_ICTPROBABAS", "SKILL_ICTCONTABAS"), 
year == 2024) |> 
left_join(countryISO, by = c("countryId" = "Code")) |>
left_join(getHRDx("skillType"), by = c("skillTypeCode" = "code")) |>
group_by(name) |>
summarize(m = symmetrize_ratio(mean(value))) |>
mutate(name = trimws(str_remove_all(name, "ICT Skill: Above basic|skills")), 
col = case_when(between(m, 1 - .01, 1.01) ~ "P", 
m < 1 - .01  ~ "M", 
.default = "F"))
```

```{r}
basic |>
ggplot(aes(x = m, y = fct_reorder(str_wrap(name, 20), m))) +
geom_point(aes(color = col), size = 5) + 
scale_color_ghro("gender") +
        geom_vline(xintercept = 1, color = "gray30", linetype = "dashed") +

scale_x_continuous(NULL, expand = c(0, 0), limits = c(0, 2), labels = function(x) {
        ifelse(
            x == 1,
            "Gender parity = 1",
            x
        )
    }) + 
    scale_y_discrete(NULL) +
annotate("richtext", x = 1.05, y = 5.5,
           label = "higher rate of →<br>basic skills<br>among <span style='color:#990099'>women</span>",
           hjust = 0, fill = NA, label.color = NA) +
annotate("richtext", x = .95, y = 5.5,
           label = "← higher rate of<br>basic skills<br>among <span style='color:#009900'>men</span>",
           hjust = 1, fill = NA, label.color = NA) +
    theme_minimal_grid() + 
    theme(legend.position = "none", plot.margin = margin(t = 20))  +
    coord_cartesian(clip = "off")
```

1. Most skills are close to parity - but info and data literacy lean towards men, whereas content creation leans towards women

```{r}
above_basic |>
ggplot(aes(x = m, y = fct_reorder(str_wrap(name, 20), m))) +
geom_point(aes(color = col), size = 5) + 
scale_color_ghro("gender") +
        geom_vline(xintercept = 1, color = "gray30", linetype = "dashed") +

scale_x_continuous(NULL, expand = c(0, 0), limits = c(0, 2), labels = function(x) {
        ifelse(
            x == 1,
            "Gender parity = 1",
            x
        )
    }) + 
    scale_y_discrete(NULL) +
annotate("richtext", x = 1.05, y = 5.5,
           label = "higher rate of →<br>above basic skills<br>among <span style='color:#990099'>women</span>",
           hjust = 0, fill = NA, label.color = NA) +
annotate("richtext", x = .95, y = 5.5,
           label = "← higher rate of<br>above basic skills<br>among <span style='color:#009900'>men</span>",
           hjust = 1, fill = NA, label.color = NA) +
    theme_minimal_grid() + 
    theme(legend.position = "none", plot.margin = margin(t = 20)) +
    coord_cartesian(clip = "off")
```

## Education and training participation

```{r}
#| include: false
#| echo: false

edutrain <- right_raw_data[['SE_GPI_PART']] |>
mutate(year = year(datetime))

dim(edutrain)
colnames(edutrain) # ageCode - choose Y15T64, most comprehensive/total
length(unique(edutrain$countryId)) # we have data for 167 countries

table(edutrain$ageCode)

unique(edutrain$regionId) # and no regions

sort(unique(year(edutrain$datetime))) # data for 2010-2025

edutrain |> count(year) # not a lot of data for 2025 - exclude it
```

```{r}
edutrain |>
filter(ageCode == "Y15T64", year != 2025) |>
group_by(year) |>
summarize(m = symmetrize_ratio(mean(value))) |>
ggplot(aes(x = year, y = m)) + 
        geom_hline(yintercept = 1, color = "gray30", linetype = "dashed") +

geom_line(color = "#006fb7") + 
scale_y_continuous(NULL, expand = c(0, 0), limits = c(0, 2), labels = function(x) {
        ifelse(
            x == 1,
            "Higher\nparticipation rate\namong women ↑\n\nGender parity = 1\n\nHigher\nparticipation rate\namong men ↓",
            x
        )
    }) + 
    scale_x_continuous(NULL, expand = c(0, 0)) +
    theme_minimal_grid()
```

1. Rate is near parity, but leans towards women participating more

## Functional training

Most recent data is from 2017, excluding this bc that's almost 10 years old. Doesn't really say a lot about currently.
```{r}
#| include: false
#| echo: false
func <- right_raw_data[["SE_ADT_FUNS"]]

dim(func)
colnames(func) 
length(unique(func$countryId))

unique(func$regionId) # and no regions

sort(unique(year(func$datetime))) # data for 2010-2017 - don't use this
```

## Reading and math
Most recent global data is from 2019 - might exclude, or use averages of country data
```{r}
#| include: false
#| echo: false
rm <- right_raw_data[["SE_TOT_PRFL"]] |>
mutate(year = year(datetime))

dim(rm)
colnames(rm) # skillTypeCode, sexCode
length(unique(rm$countryId)) # we have data for 147 countries

unique(rm$regionId) # and a few regions

sort(unique(year(rm$datetime))) # data for 2010-2024
```

```{r}
# world_rm <- rm |> 
# filter(regionId == 1, sexCode == "_T") |>
# left_join(edu, by = c("educationCode" = "code"))


# pct_change <- function(new, old) {
#   (new - old)/abs(old)
# }

# p_math_change <- pct_change(p_math_2019, p_math_2010)
# p_read_change <- pct_change(p_read_2019, p_read_2010)
# ls_math_change <- pct_change(ls_math_2019, ls_math_2010)
# ls_read_change <- pct_change(ls_read_2019, ls_read_2010)

# world_rm <- world_rm |> 
# mutate(annotation = case_when(year == 2019 & skillTypeCode == "SKILL_MATH" & name == "Lower secondary education" ~ ls_math_change, 
# year == 2019 & skillTypeCode == "SKILL_MATH" & name == "Primary education" ~ p_math_change, 
# year == 2019 & skillTypeCode == "SKILL_READ" & name == "Primary education" ~ p_read_change, 
# year == 2019 & skillTypeCode == "SKILL_READ" & name == "Lower secondary education" ~ ls_read_change, 
# .default = NA), 
# annotation = ifelse(annotation > 0, 
# paste0(scales::percent(annotation, accuracy = .1), " increase in\n", ifelse(skillTypeCode == "SKILL_READ", "reading proficiency rate", "mathematics proficiency rate")), 
# paste0(scales::percent(abs(annotation), accuracy = .1), " decrease in\n", ifelse(skillTypeCode == "SKILL_READ", "reading proficiency rate", "mathematics proficiency rate"))))

# world_rm |>
# ggplot(aes(x = year, y = value)) + 
# geom_line(aes(color = skillTypeCode)) + 
# geom_text(aes(label = annotation, color = skillTypeCode), hjust = 0) + 
# facet_wrap(~fct_rev(name), ncol = 1) +
# scale_color_manual(NULL, values = c("SKILL_MATH" = palettes$cats[3], "SKILL_READ" = palettes$cats[4]), labels = c("SKILL_MATH" = "Mathematics", "SKILL_READ" = "Reading"), guide = "none") +
# scale_y_continuous("% children and youth achieving minimum proficiency level", limits = c(0, 100), expand = c(0, 0)) +
# scale_x_continuous(NULL, expand = c(0, 0), breaks = seq(2010, 2019, by = 3)) + 
# theme_minimal_grid() +
# panel_border() + 
# theme(plot.margin = margin(r = 200)) +
# coord_cartesian(clip = "off")
```

1. Lower secondary's level of literacy and mathematics is improving slightly, whereas primary is falling

## parity read math

```{r}
g_rm <- right_raw_data[["SE_TOT_GPI"]] |> 
mutate(year = year(datetime)) |>
filter(year == max(year), educationCode == "ISCED11_10") |> 
group_by(educationCode, skillTypeCode, year) |>
summarize(m = mean(value), n = n(), code = "Gender") |> 
ungroup() |>
filter(n > 10)

lang_rm <- right_raw_data[["SE_LGP_ACHI"]] |> 
mutate(year = year(datetime)) |>
filter(year == max(year), educationCode == "ISCED11_10") |> 
group_by(educationCode, skillTypeCode, year) |>
summarize(m = mean(value), n = n(), code = "Language test") |> 
ungroup() |>
filter(n > 10)

imm_rm <- right_raw_data[["SE_NAP_ACHI"]] |> 
mutate(year = year(datetime)) |>
filter(year == max(year), educationCode == "ISCED11_10") |> 
group_by(educationCode, skillTypeCode, year) |>
summarize(m = mean(value), n = n(), code = "Immigration status") |> 
ungroup() |>
filter(n > 10)

ur_rm <- right_raw_data[["SE_TOT_RUPI"]] |> 
mutate(year = year(datetime)) |>
filter(year == max(year), educationCode == "ISCED11_10") |> 
group_by(educationCode, skillTypeCode, year) |>
summarize(m = mean(value), n = n(), code = "Rural-urban") |> 
ungroup() |>
filter(n > 10)

se_rm <- right_raw_data[["SE_TOT_SESPI"]] |> 
mutate(year = year(datetime)) |>
filter(year == max(year), educationCode == "ISCED11_10") |> 
group_by(educationCode, skillTypeCode, year) |>
summarize(m = mean(value), n = n(), code = "Socio-economic") |> 
ungroup() |>
filter(n > 10)

rm_ratios <- rbind(g_rm, lang_rm, imm_rm, ur_rm, se_rm)

rm_ratios |>
ggplot(aes(x = code, y = m)) + 
geom_point(color = "#006fb7", size = 5) +
        geom_hline(yintercept = 1, color = "gray30", linetype = "dashed") +
theme_minimal_hgrid()
```

1. Socio-economic status has the biggest gap in proficiency, followed by language test and rural-urban changes

# Organized learning
```{r}
org_learn <- right_raw_data[['SE_GPI_PTNPRE']] |> 
filter(regionId == 1) |>
mutate(year = year(datetime), value = symmetrize_ratio(value))

org_learn |> 
ggplot(aes(x = year, y = value)) + 
geom_line(color = "#006fb7") +         
geom_hline(yintercept = 1, color = "gray30", linetype = "dashed") + 
theme_minimal_grid() + 
scale_y_continuous(limits = c(.5, 1.5), expand = c(0, 0))
```

1. Very close to parity, meaning women are almost just as involved in organized learning as men

# Education and training

```{r}
# edu_train <- right_raw_data[['SE_GPI_PART']] |> 
# filter(ageCode != "Y15T64") |>
# group_by(ageCode, datetime) |>
# summarize(m = symmetrize_ratio(mean(value))) |>
# mutate(year = year(datetime))

# edu_train |>
# ggplot(aes(x = ))
```

# Qualified teachers

```{r}
#| echo: false
#| include: false

teachers <- right_raw_data[["SE_TRA_GRDL"]] |>
    mutate(year = year(datetime))

dim(teachers)
colnames(teachers) # educationCode, sexCode
length(unique(teachers$countryId)) # we have data for 165 countries

unique(teachers$regionId) # and some regions

sort(unique(year(teachers$datetime))) # data for 2010-2025
```

```{r}
teachers |>
    filter(regionId == 1, sexCode == "_T", educationCode %in% c("ISCED11_10", "ISCED11_2", "ISCED97_3"), year > 2015) |>
    left_join(edu, by = c("educationCode" = "code")) |>
    ggplot(aes(x = year, y = value)) +
    geom_line(aes(color = factor(name, levels = c("Primary education", "Lower secondary education", "Upper secondary education")))) +
    scale_color_manual("Education level", values = c("Primary education" = pal[4], "Lower secondary education" = pal[7], "Upper secondary education" = pal[11])) +
    scale_y_continuous("% teachers") +
    theme_minimal_grid()
```