---
title: "Right to food"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
execute:
  warning: false
  message: false
  out-width: 100%
---

```{r setup}
#| echo: false
library(magrittr)
library(showtext)
library(hrbrthemes)
library(waffle)
library(tidyverse)
library(cowplot)
library(rvest)
library(ggbeeswarm)
library(scales)
library(httr)
library(jsonlite)
library(ggrepel)
library(readxl)
library(ggmosaic)
library(ggarrow)
library(here)
library(ggtext)
source(here("ghro.R"))

raw <- here("data/raw/")

dir.create(here("drafts/rightToFood/datasets"), recursive = TRUE, showWarnings = FALSE)
```

Indicators

- Prevalence of undernourishment (45)

  - number of undernourished people (SN_ITK_DEFCN)

  - prevalence of undernourishment (SN_ITK_DEFC)

- Prevalence of moderate or severe food insecurity... (47)

  - Severe food insecurity % (AG_PRD_FIESS)

  - Severe food insecurity n (AG_PRD_FIESSN)

  - Moderate or severe food insecurity n (AG_PRD_FIESMSN)

  - Moderate or severe food insecurity % (AG_PRD_FIESMS)

- Prevalence of malnutrition (49)

  - Children moderately or severely wasted n (SH_STA_WASTN)

  - Children moderately or severely wasted % (SH_STA_WAST)

  - Children moderately or severely overweight n (SN_STA_OVWGTN)

  - Children moderately or severely overweight % (SN_STA_OVWGT)

Note: I think it would be helpful to explain the differences between all of these terms

- Malnutrition: deficiencies, excesses, or imbalances in a person’s intake of energy and/or nutrients.

- Food insecurity: when people lack secure access to sufficient amounts of safe and nutritious food for normal growth and development and an active and healthy life.

- Undernourishment: people whose dietary energy consumption is continuously below a minimum dietary energy requirement for maintaining a healthy life and carrying out a light physical activity

```{r}
#| echo: false
#| include: false
codes <- c("SN_STA_OVWGT", "SN_STA_OVWGTN", "SH_STA_WAST", "SH_STA_WASTN", "AG_PRD_FIESMSN", "AG_PRD_FIESMS", "AG_PRD_FIESSN", "SN_ITK_DEFC", "AG_PRD_FIESS", "SN_ITK_DEFCN")

datasets <- list()

for (code in codes) {
    ds <- getHRDx(code, offline = TRUE, raw = here("data/raw/rightToFood"))

    datasets[[code]] <- ds

    #write.csv(ds, here(paste0("data/raw/rightToFood/", code, ".csv")))
}
```

# Undernourishment

```{r}
#| echo: false
#| include: false
undernourish_prev <- datasets[["SN_ITK_DEFC"]] |>
mutate(year = year(datetime))

colnames(undernourish_prev) # no disaggs ??

table(year(undernourish_prev$datetime)) # data coverage declines in 2024

table(undernourish_prev$regionId) # but there are global figures already in the data
```

```{r undernourish-world}
undernourish_prev |> 
filter(regionId == 1) |>
ggplot(aes(x = year, y = value)) + 
geom_line(color = "#006fb7") + 
scale_y_continuous("Prevalence of undernourishment", limits = c(0, 15), 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_x_continuous(NULL) + 
theme_minimal_grid()
```

```{r undernourish-regions}
main <- c("Africa", "Asia", "Americas", "Europe", "Oceania")
undernourish_prev |>
left_join(regions |> drop_na(), by = c("regionId" = "id")) |>
filter(level == 1) |> 
ggplot(aes(x = year, y = value)) +
geom_line(aes(color = name)) + 
scale_color_ghro("region") + 
scale_y_continuous("Prevalence of undernourishment", limits = c(0, 25), 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_x_continuous(NULL) +  
theme_minimal_grid()
```

1.  No data for Europe?
2.  Asia saw a jump in 2020, but is declining now
3.  Africa is on a concerning increase, reaching above 20%, a new high since 2010

```{r}
#| echo: false
#| include: false
undernourish_num <- datasets[["SN_ITK_DEFCN"]] |>
mutate(year = year(datetime))

eda(undernourish_num, detailed = TRUE)
```

```{r undernourish-n-world}
undernourish_num |>
filter(regionId == 1) |>
arrange(year) |>
ggplot(aes(x = year)) + 
geom_line(aes(y = value/1000000), color = "#006fb7") +
scale_y_continuous("Number undernourished, millions of people",  
                   expand = c(0, 0)) + 
scale_x_continuous(NULL) + 
theme_minimal_grid()

# prop undernourished globally, manual calculation w unpop numbers
pop2024 <- read.csv(here(raw, "/worldpop.csv")) |> filter(Time == 2024) |> pull(Value)

undernourish2024 <- undernourish_num |> filter(year == 2024, regionId == 1) |> pull(value)

undernourish2024/pop2024
```

1.  673.2 million people were undernourished in 2024.
2.  This means that about 1 in a dozen people globally do not have access to food sufficient to maintain a healthy life.
3.  Worst was 2021, but slowly declining in post-pandemic era

# Food insecurity

```{r sev-fi-world}
sev <- datasets[["AG_PRD_FIESS"]] |> mutate(year = year(datetime))
modsev <- datasets[["AG_PRD_FIESMS"]] |> mutate(year = year(datetime))

sev_world <- sev |> filter(regionId == 1, ageCode == "_T", dwellingUrbanisationCode == "_T")

sev_world |> 
ggplot(aes(x = year, y = value)) + 
geom_line(color = "#006fb7") +
scale_y_continuous("Prevalence of severe food insecurity", 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_x_continuous(NULL) + 
theme_minimal_grid()
```

1.  Same trend with pandemic - still at all time high
2.  1/10 of world in severe

```{r modsev-fi-world}
modsev_world <- modsev |> filter(regionId == 1, ageCode == "_T", dwellingUrbanisationCode == "_T")

modsev_world |>
ggplot(aes(x = year, y = value)) + 
geom_line(color = "#006fb7") +
scale_y_continuous("Prevalence of moderate or severe food insecurity", 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_x_continuous(NULL) + 
theme_minimal_grid()
```

Maybe show them together?

Reference for showing just moderate: <https://www.fao.org/interactive/state-of-food-security-nutrition/en/>

```{r fi-world}
#| fig-width: 8
#| fig-height: 6
food_insecurity_world <- rbind(modsev_world |> mutate(type = "Moderate or severe"), sev_world |> mutate(type = "Severe")) |> select(value, year, type) |> pivot_wider(names_from = type, values_from = value) |> 
mutate(Moderate = `Moderate or severe` - Severe) |>
pivot_longer(c("Severe", "Moderate")) |>
mutate(lab = ifelse(name == "Moderate", `Moderate or severe`, value))

food_insecurity_world |>
ggplot(aes(x = year, y = value)) + 
geom_col(aes(fill = name)) +
scale_fill_manual("Level of food insecurity", values = c("Severe" = "#D7191C", "Moderate" = "#FDAE61")) +
geom_text(aes(y = lab - 1, label = scales::percent(value, accuracy = .1, scale = 1), color = name), size = 4) +
scale_color_manual(values = c("Severe" = "white", "Moderate" = "black"), guide = "none") +
scale_y_continuous("Prevalence of food insecurity", 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_x_continuous(NULL) + 
theme_minimal_hgrid() + 
theme(legend.position = "top")
```

1.  While prevalence of moderate or severe food insecurity is lower than it's been since 2019, it is still much higher than it's been in this indicator's history.

```{r sev-fi-regions}
sev_regions <- sev |> 
filter(!is.na(regionId), ageCode == "_T", dwellingUrbanisationCode == "_T") |>
left_join(regions, by = c("regionId" = "id"))

labs <- sev_regions |> filter(year == 2024, level == 1) 

sev_regions |> 
filter(level == 1) |> 
ggplot(aes(x = year, y = value)) +
geom_line(aes(color = name)) +
ggrepel::geom_text_repel(data = labs, aes(y = value, label = paste0(name, "\n", scales::percent(value, scale = 1, accuracy = .1)), color = name), hjust = 0, xlim = c(2024, Inf)) +
scale_color_ghro("region") +
guides(color = "none") + scale_y_continuous("Prevalence of severe food insecurity", 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_x_continuous(NULL) + 
theme_minimal_grid() +
theme(plot.margin = margin(r = 40)) +
coord_cartesian(clip = "off")
```

1.  Breaking it down by region shows that the decline in severe food insecurity is driven by decreases in the Americas and Asia.
2.  However, people in Africa are more than twice as likely to be in severe food insecurity compared to any other region.
3.  This trend does not seem to be improving, and is instead worsening, reaching 22.7% in 2024.

```{r sev-fi-africa}
sev_africa <- sev_regions |> filter(parentId %in% c(2, 202), year == 2024, name != "Sub-Saharan Africa")

sev_africa |> 
ggplot(aes(y = fct_reorder(name, value), x = value)) + 
geom_col(fill = "#006fb7") + 
scale_x_continuous("Prevalence of severe food insecurity, 2024", 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_y_discrete(NULL) + 
theme_minimal_vgrid() + 
theme(plot.margin = margin(l = 50)) 
```

1.  Looking at sub-regions of Africa, this trend is largely driven by Middle Africa, where nearly 2 in 5 people live in severe food insecurity.

```{r modsev-fi-regions}
modsev_regions <- modsev |> 
filter(!is.na(regionId), ageCode == "_T", dwellingUrbanisationCode == "_T") |>
left_join(regions, by = c("regionId" = "id")) |> 
mutate(year = year(datetime))

labs <- modsev_regions |> filter(year == 2024, level == 1) 

modsev_regions |> 
filter(level == 1) |> 
ggplot(aes(x = year, y = value)) +
geom_line(aes(color = name)) +
ggrepel::geom_text_repel(data = labs, aes(y = value, label = paste0(name, "\n", scales::percent(value, scale = 1, accuracy = .1)), color = name), hjust = 0, xlim = c(2024, Inf)) +
scale_color_ghro("region") +
guides(color = "none") + scale_y_continuous("Prevalence of moderate or severe food insecurity", 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_x_continuous(NULL) + 
theme_minimal_grid() +
theme(plot.margin = margin(r = 40)) +
coord_cartesian(clip = "off")
```

1.  In Africa, 3 in 5 live in moderate or severe food insecurity

```{r loc-fi}
du_sev <- sev |> filter(dwellingUrbanisationCode != "_T", regionId == 1, year == 2024) |>
mutate(type = "Severe")

du_modsev <- modsev |> filter(dwellingUrbanisationCode != "_T", regionId == 1, year == 2024) |>
mutate(type = "Moderate or severe")

key <- getHRDx("dwellingUrbanisation") |> select(code, name) |> rename(dwellingUrbanisationCode = code, location = name)

food_insecurity_du <- rbind(du_sev, du_modsev) |> select(value, type, dwellingUrbanisationCode) |> pivot_wider(names_from = type, values_from = value) |> 
mutate(Moderate = `Moderate or severe` - Severe) |>
pivot_longer(c("Severe", "Moderate")) |>
mutate(lab = ifelse(name == "Moderate", `Moderate or severe`, value)) |>
left_join(key, by = "dwellingUrbanisationCode")

food_insecurity_du |>
ggplot(aes(y = fct_reorder(str_wrap(location, 25), value), x = value)) + 
geom_col(aes(fill = name)) +
scale_fill_manual("Level of food insecurity", values = c("Severe" = "#D7191C", "Moderate" = "#FDAE61")) +
geom_text(aes(x = lab - .1, label = scales::percent(value, accuracy = .1, scale = 1), color = name), size = 4, hjust = 1) +
scale_color_manual(values = c("Severe" = "white", "Moderate" = "black"), guide = "none") +
scale_x_continuous("Prevalence of food insecurity among those 15 and older", 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_y_discrete(NULL) + 
theme_minimal_vgrid() + 
theme(legend.position = "top") 
```

1.  Food insecurity is most prevalent in rural areas

```{r sex-fi}
sev_sex <- sev |> 
filter(regionId == 1, sexCode != "_T", year == 2024) |>
mutate(type = "Severe")

modsev_sex <- modsev |> 
filter(regionId == 1, sexCode != "_T", year == 2024) |>
mutate(type = "Moderate or severe")

food_insecurity_sex <- rbind(modsev_sex, sev_sex) |> select(value, year, type, sexCode) |> pivot_wider(names_from = type, values_from = value) |> 
mutate(Moderate = `Moderate or severe` - Severe) |>
pivot_longer(c("Severe", "Moderate"), names_to = "type") |>
mutate(lab = ifelse(type == "Moderate", `Moderate or severe`, value))

food_insecurity_sex |> 
ggplot(aes(y = fct_reorder(str_wrap(sexCode, 25), value), x = value)) + 
geom_col(aes(fill = type)) +
scale_fill_manual("Level of food insecurity", values = c("Severe" = "#D7191C", "Moderate" = "#FDAE61")) +
geom_text(aes(x = lab - .1, label = scales::percent(value, accuracy = .1, scale = 1), color = type), size = 4, hjust = 1) +
scale_color_manual(values = c("Severe" = "white", "Moderate" = "black"), guide = "none") +
scale_x_continuous("Prevalence of food insecurity among those 15 and older", 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_y_discrete(NULL, labels = c("F" = "Female", "M" = "Male")) + 
theme_minimal_vgrid() + 
theme(legend.position = "top") 
```

# Malnutrition

```{r}
wasting <- datasets[["SH_STA_WAST"]] |>
mutate(year = year(datetime))

wasting_world <- wasting |>
filter(regionId == 1)
```

```{r wasting-world}
wasting_world |>
ggplot(aes(x = year, y = value)) + 
geom_line(color = "#006fb7") +
scale_y_continuous("Prevalence of wasting among children under 5",  
                   labels = 
             scales::percent_format(scale = 1), 
                   limits = c(0, 15), 
                   expand = c(0, 0)) + 
scale_x_continuous(NULL) + 
theme_minimal_grid()
```

1.  Globally since 2010, improvement of 1.1% (not a huge change)

```{r wasting-regions}
wasting_regions <- wasting |> filter(!is.na(regionId)) |> left_join(regions, by = c("regionId" = "id")) |> filter(level == 2) |> left_join(regions |> select(id, name) |> rename(parent_name = name), by = c("parentId" = "id"))

wasting_regions |> 
filter(year == max(year)) |> 
ggplot(aes(y = fct_reorder(name, value), x = value)) +
geom_col(aes(fill = parent_name)) + 
geom_text(aes(x = value + .1, label = scales::percent(value, accuracy = .1, scale = 1)), size = 3.5, hjust = 0) +
scale_fill_ghro("region") + 
scale_x_continuous("Prevalence of wasting among children under 5, 2024", 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_y_discrete(NULL) + 
theme_minimal_vgrid() + 
theme(legend.position = "top", plot.margin = margin(l = 30, r = 30)) + 
coord_cartesian(clip = "off")
```

1.  regional values range from .2 in northern america, to 13.6% in southern asia (nearly double South-eastern Asia, the next highest regional value), but this is on a consistent downwards trend
2.  No data for a good amount of Oceania, hence only Australia and New Zealand

```{r, include = FALSE}
wasting_regions |> filter(name == "Southern Asia") |> select(year, value) |> arrange(year)

wasting |> 
filter(is.na(regionId)) |>
left_join(countries, by = c("countryId" = "id")) |> filter(level1 == 9) |> filter(year > 2022)
```

```{r overwt-world}
overwt <- datasets[["SN_STA_OVWGT"]] |> mutate(year = year(datetime))

overwt_world <- overwt |> filter(regionId == 1, sexCode == "_T")

overwt_world |>
ggplot(aes(x = year, y = value)) + 
geom_line(color = "#006fb7") +
scale_y_continuous("Prevalence of overweight among children under 5",  
                   labels = 
             scales::percent_format(scale = 1), 
                   limits = c(0, 10), 
                   expand = c(0, 0)) + 
scale_x_continuous(NULL) + 
theme_minimal_grid()
```

1.  Mostly stagnant, but small increase since 2019 (2024 = highest since 2010)

```{r overwt-regions}
#| fig-height: 6
#| fig-width: 7
overwt_regions <- overwt |> filter(!is.na(regionId), year == 2024, sexCode == "_T") |> left_join(regions |> select(id, name, level, parentId), by = c("regionId" = "id")) |> filter(level == 2) |> left_join(regions |> select(id, name) |> rename(parent_name = name), by = c("parentId" = "id"))

global_val <- overwt_world |> filter(year == 2024) |> pull(value)

overwt_regions |> ggplot(aes(y = fct_reorder(str_wrap(name, 20), value), x = value)) +
geom_col(aes(fill = parent_name)) + 
geom_text(aes(x = value + .1, label = scales::percent(value, accuracy = .1, scale = 1)), size = 3.5, hjust = 0) +
geom_vline(xintercept = global_val, linetype = "dashed", color = "grey40") + 
scale_fill_ghro("region") + 
scale_x_continuous("Prevalence of overweight among children under 5, 2024", 
                   labels = scales::percent_format(scale = 1), 
                   expand = c(0, 0)) + 
scale_y_discrete(NULL) + 
theme_minimal_vgrid() + 
theme(legend.position = "top", plot.margin = margin(l = 30, r = 30)) + 
coord_cartesian(clip = "off")
```