---
title: "Right to development"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
execute:
  warning: false
  message: false
  out-width: 100%
---

```{r setup}
#| echo: false
library(magrittr)
library(showtext)
library(hrbrthemes)
library(waffle)
library(tidyverse)
library(cowplot)
library(rvest)
library(ggbeeswarm)
library(scales)
library(httr)
library(jsonlite)
library(ggrepel)
library(readxl)
library(ggmosaic)
library(ggarrow)
library(here)
library(ggtext)
source(here("consts.R"))

dir.create(here("drafts/rightToDevelopment/datasets"), recursive = TRUE, showWarnings = FALSE)
```

```{r}
url <- "https://sdmx.oecd.org/public/rest/data/OECD.WISE.RSB,DSD_SDG@DF_SDG_G_17,2.0/..17_2..._T._T._T._T._T.?startPeriod=2020&dimensionAtObservation=AllDimensions&format=csvfilewithlabels"

sdg1721 <- read.csv(url)
```


```{r}
#| fig-height: 10
#| fig-width: 7
#| out-width: 100%
codes <- c("17_2_1_DC_ODA_TOTGGE", "17_2_1_DC_ODA_LDCG")

sdg1721 |>
    filter(SDG_SERIES == codes[1], TIME_PERIOD == 2024) |>
    mutate(color = ifelse(OBS_VALUE > .7, "Y", "N")) |>
    ggplot(aes(y = fct_reorder(Reference.area, OBS_VALUE), x = OBS_VALUE)) +
    geom_vline(xintercept = .7, color = "gray30", linetype = "dashed", alpha = .7) +
    geom_col(aes(fill = color)) +
    geom_text(
        aes(
            label = scales::percent(OBS_VALUE, scale = 1, accuracy = .01),
            x = OBS_VALUE + .01
        ),
        hjust = 0
    ) +
    scale_fill_manual(values = c("Y" = "#006fb7", "N" = colorspace::lighten("#006fb7", .8)), guide = "none") +
    scale_x_continuous(NULL, expand = c(0, 0), labels = NULL, breaks = NULL) +
    scale_y_discrete(NULL) +
    theme_minimal_vgrid() +
    theme(plot.margin = margin(r = 75), axis.ticks.x = element_blank()) +
    coord_cartesian(clip = "off")
```

```{r}
sdg1721 |>
    filter(SDG_SERIES == codes[1]) |>
    group_by(TIME_PERIOD) |>
    summarize(m = mean(OBS_VALUE)) |>
    ggplot(aes(x = TIME_PERIOD, y = m)) +
    geom_line(color = "#006fb7") +
    scale_x_continuous(NULL, expand = c(0, 0)) +
    scale_y_continuous("average ODA on grant equivalent basis (% GNI)") + 
    theme_minimal_grid() +
    theme(plot.margin = margin(r = 20))
```

```{r}
url2 <- "https://sdmx.oecd.org/public/rest/data/OECD.DCD.FSD,DSD_DAC2@DF_DAC2A,/DAC.LDC+ALLR.206.USD.Q?startPeriod=2020&dimensionAtObservation=AllDimensions&format=csvfilewithlabels"

disburse <- read.csv(url2)

dac <- disburse |>
    arrange(Recipient, TIME_PERIOD) |>
    group_by(Recipient) |>
    mutate(
        lag = lag(OBS_VALUE),
        pct_change = (OBS_VALUE - lag) / lag
    )
```

```{r}
dac |>
    ggplot(aes(x = TIME_PERIOD, y = OBS_VALUE)) +
    geom_line(aes(color = Recipient)) +
    scale_color_ghro("misc") +
    scale_x_continuous(NULL) +
    scale_y_continuous("Millions of US dollars, 2024", labels = scales::comma_format()) +
    theme_minimal_grid()
```

1. Disbursements to LDCs have been relatively constant, but have been on a slow decline the past few years, whereas 2024 is the first year that it decreased for all recipients/overall. 

Also the dataset detailed in the GHRIF should reflect that it is now measured on grant equivalent basis, which isn't super clear from what it currently says?


reduction of ODA vs increase in defense = https://www.sipri.org/databases/milex 

mention the names of countries that meet target, do not mention those that do not

cat of ODA that goes to human rights institutions - some on gender, refugees, etc. 

look at before vs after taxes and social stuff
free economy vs govt interference (you expect inequality to be reduced w govt interference)

# Gini inequality

```{r}
gini <- read.csv(here("data/raw/gini.csv"), nrow = 265)

gini2 <- gini |>
    pivot_longer(starts_with("X")) |>
    mutate(
        name = as.integer(str_sub(name, 2, 6)),
        value = replace(value, value == "..", NA)
    ) |>
    drop_na() |>
    group_by(Country.Name) |>
    filter(name == max(name))

bands <- data.frame(
    level = factor(c("Low", "Moderately low", "Moderately high", "High", "Very high"), levels = c("Low", "Moderately low", "Moderately high", "High", "Very high")),
    min = c(-Inf, 25, 30, 40, 45),
    max = c(25, 30, 40, 45, Inf)
)

pal <- dichromat::colorschemes$DarkRedtoBlue.12[c(2, 4, 8, 10, 11)]
names(pal) <- bands$level

g1 <- gini2 |>
    filter(name > 2015) |>
    group_by(name) |>
    summarize(
        m = mean(as.numeric(value)),
        n_countries = n()
    ) |>
    ggplot(aes(x = name, y = m)) +
    geom_rect(data = bands, aes(ymin = min, ymax = max, fill = level), xmax = Inf, xmin = -Inf, alpha = .5, inherit.aes = FALSE) +
    scale_fill_manual(values = pal) +
    geom_line() +
    scale_x_continuous(NULL, expand = c(0, 0), breaks = seq(2016, 2025, by = 2)) +
    scale_y_continuous("Gini index", limits = c(0, 55), expand = c(0, 0)) +
    theme_minimal_grid() +
    coord_cartesian(clip = "off") +
    theme(plot.margin = margin(r = 10, t = 5), legend.position = "none") +
    ggtitle("Non-weighted gini,\nnot filled")
```

I feel like there is a more intuitive way to show this, because currently it shows it increasing which looks to be good

Notes: 
- SDG goal is below 30
- high inequality is above 40

```{r}
#| fig-width: 10
#| fig-height: 6
library(patchwork)
pop <- read.csv(here("data/raw/unpop2000.csv")) |> select(Iso3, Time, Value)

test <- gini2 |>
    ungroup() |>
    mutate(name = as.integer(name)) |>
    complete(nesting(Country.Name, Country.Code, Series.Name, Series.Code), name = 2000:2025) |>
    arrange(Country.Name, name) |>
    group_by(Country.Name) |>
    fill(value, .direction = "down") |>
    ungroup() |>
    left_join(pop, by = c("Country.Code" = "Iso3", "name" = "Time"))

weighted_gini_by_year <- test %>%
    mutate(value = as.numeric(value)) %>%
    filter(!is.na(value)) %>%
    group_by(name) %>%
    summarise(
        weighted_gini = sum(value * Value) / sum(Value),
        n_countries   = n()
    )

weighted_gini_by_year

g2 <- weighted_gini_by_year |>
    filter(name > 2015) |>
    ggplot(aes(x = name, y = weighted_gini)) +
    geom_rect(data = bands, aes(ymin = min, ymax = max, fill = level), xmax = Inf, xmin = -Inf, alpha = .5, inherit.aes = FALSE) +
    scale_fill_manual(values = pal) +
    geom_line() +
    scale_x_continuous(NULL, expand = c(0, 0), breaks = seq(2016, 2025, by = 2)) +
    scale_y_continuous("Gini index", limits = c(0, 55), expand = c(0, 0)) +
    theme_minimal_grid() +
    coord_cartesian(clip = "off") +
    theme(plot.margin = margin(r = 10, t = 5), legend.position = "none") +
    ggtitle("Weighted gini,\nfilled from 2020 on")

g3 <- test %>%
    mutate(value = as.numeric(value)) %>%
    filter(name > 2015) |>
    filter(!is.na(value)) %>%
    group_by(name) %>%
    summarise(
        m = mean(value),
        n_countries = n()
    ) |>
    ggplot(aes(x = name, y = m)) +
    geom_rect(data = bands, aes(ymin = min, ymax = max, fill = level), xmax = Inf, xmin = -Inf, alpha = .5, inherit.aes = FALSE) +
    scale_fill_manual(values = pal) +
    geom_line() +
    scale_x_continuous(NULL, expand = c(0, 0), breaks = seq(2016, 2025, by = 2)) +
    scale_y_continuous("Gini index", limits = c(0, 55), expand = c(0, 0)) +
    theme_minimal_grid() +
    coord_cartesian(clip = "off") +
    theme(plot.margin = margin(r = 10, t = 5), legend.position = "none") +
    ggtitle("Non-weighted gini,\nfilled from 2020 on")

g1 + g2 + g3
```

```{r}
test2 <- gini2 |>
    ungroup() |>
    mutate(name = as.integer(name)) |>
    complete(nesting(Country.Name, Country.Code, Series.Name, Series.Code), name = 2000:2025) |>
    arrange(Country.Name, name) |>
    group_by(Country.Name) |>
    fill(value, .direction = "down") |>
    ungroup() |>
    mutate(high_ineq = ifelse(as.numeric(value) >= 40, 1, 0))

#test2 |> distinct(Country.Name, value)
```

### gini but from the sdg database

```{r}
sdg_gini <- readxl::read_excel(here("data/raw/SI_DST_FISP.xlsx")) |>
    select(-SeriesCode, -Goal, -Target, -Indicator, -SeriesDescription) |>
    pivot_longer(cols = `2007`:`2025`)

sdg_gini_avg <- sdg_gini |>
    group_by(name, `Fiscal intervention stage`) |>
    drop_na(value) |>
    filter(value != NaN, !name %in% c(2007, 2025, 2013)) |>
    summarize(
        m = mean(as.numeric(value)),
        n = n()
    )

sdg_gini_avg |>
    filter(`Fiscal intervention stage` != "POSTFIS_CON_INC") |>
    ggplot(aes(x = as.integer(name), y = m)) +
    geom_rect(data = bands, aes(ymin = min, ymax = max, fill = level), xmax = Inf, xmin = -Inf, alpha = .5, inherit.aes = FALSE) +
    scale_fill_manual(values = pal) +
    geom_line(aes(group = `Fiscal intervention stage`)) +
    annotate("text", x = 2024, y = c(31.25, 41.4), label = c("Post-fiscal\ndisposable", "Pre-fiscal"), hjust = 0) +
    annotate("text", x = 2014, y = c(20, 25, 30, 40, 45) + .1, label = c("Low", "Moderately low", "Moderately high", "High", "Very high"), color = colorspace::darken(pal, 0.3), vjust = 0, hjust = 0) +
    scale_x_continuous(NULL, expand = c(0, 0), breaks = seq(2014, 2024, by = 2)) +
    scale_y_continuous("Gini index", limits = c(20, 50), expand = c(0, 0)) +
    theme_minimal_grid() +
    theme(plot.margin = margin(r = 75, t = 5), legend.position = "none") +
coord_cartesian(clip = "off") 

sdg_gini_avg |>
    select(name, `Fiscal intervention stage`, n) |>
    pivot_wider(names_from = `Fiscal intervention stage`, values_from = n)
```

```{r}
sdg_gini |>
drop_na(value) |>
filter(`Fiscal intervention stage` != "POSTFIS_CON_INC", !name %in% c(2007, 2013, 2025)) |>
ggplot(aes(x = as.factor(name), y = as.numeric(value))) + 
geom_boxplot() +
facet_wrap(~factor(`Fiscal intervention stage`, levels = c("PREFIS_INC", "POSTFIS_DIS_INC"), labels = c("Pre-fiscal", "Post-fiscal disposable"))) + 
cowplot::theme_minimal_grid() + 
panel_border() + 
labs(x = NULL, y = "Gini index", title = "Distribution of Gini index in countries")
```

```{r}
library(httr)
library(jsonlite)
library(dplyr)

base_url <- "https://data360api.worldbank.org/data360/data"

getWB <- function(indicator) {
    database <- "WB_WDI"
    page_size <- 1000

    all_data <- list()
    skip <- 0
    i <- 1

    repeat {
        resp <- GET(base_url, query = list(
            DATABASE_ID = database,
            INDICATOR   = indicator,
            skip        = skip
        ))

        stop_for_status(resp)

        parsed <- fromJSON(content(resp, as = "text", encoding = "UTF-8"), flatten = TRUE)

        # The records are typically under a "value" field — check parsed$value or similar
        # Adjust this depending on the actual JSON structure returned
        records <- parsed$value

        if (is.null(records) || length(records) == 0 || nrow(records) == 0) {
            break
        }

        all_data[[i]] <- records
        n_returned <- nrow(records)

        cat("Fetched", n_returned, "records\n")

        # Stop when we get less than a full page (last page)
        if (n_returned < page_size) {
            break
        }

        skip <- skip + page_size
        i <- i + 1
    }
    bind_rows(all_data)
}

lowest10 <- getWB("WB_WDI_SI_DST_FRST_10")

lowest10$measure <- "Lowest 10%"

highest10 <- getWB("WB_WDI_SI_DST_10TH_10")

highest10$measure <- "Highest 10%"

compare_deciles <- rbind(lowest10, highest10)

# making sure there's no disaggs
# compare_deciles |>
#     group_by(TIME_PERIOD, measure) |>
#     count(REF_AREA) |>
#     filter(n > 1)

deciles <- compare_deciles |>
filter(between(as.integer(TIME_PERIOD), 2010, 2024)) |>
    group_by(TIME_PERIOD, measure) |>
    summarize(n = n(), 
    m = mean(as.numeric(OBS_VALUE)))
    
    deciles |>
    ggplot(aes(x = as.integer(TIME_PERIOD), y = m)) + 
    geom_line(aes(color = measure)) +
    scale_color_ghro("misc") +
    theme_minimal_grid() + 
    labs(x = NULL, y = "% of income held by population decile")
```